### 2023-01-22 paper.tex
Add initial materials

\documentclass[12pt]{article}

\usepackage{caption}
\usepackage{graphicx}
\usepackage{tipa}
\usepackage{ot-tableau}
\usepackage[margin=1in]{geometry}

\DeclareCaptionLabelFormat{parensnum}{(#2)}
\captionsetup[figure]{labelformat=parensnum,
                      labelsep=space,
                      justification=raggedright,
                      singlelinecheck=false}

\newcommand{\ident}{\textsc{Ident}}

\newcommand{\maxvplus}{\textsc{MaxFinalV}}
\newcommand{\maxplusv}{\textsc{MaxInitialV}}
\newcommand{\maxc}{\textsc{Max}}
\newcommand{\ftbin}{\textsc{FtBin}}
\newcommand{\ssp}{\textsc{SSP}}
\newcommand{\dep}{\textsc{Dep}}
\newcommand{\onset}{\textsc{Onset}}
\newcommand{\nocoda}{\textsc{NoCoda}}

\newcommand{\pref}[1]{(\ref{#1})}

\title{Vowel Elision in Yoruba Collocations}
\author{Hemant Gouni}

\begin{document}

\maketitle

In Yoruba, deletion occurs at morpheme boundaries. Specifically, when two
vowels become adjacent via concatenation of Yoruba morphemes, the first vowel
is generally deleted. This is shown in the following pairs of underlying and
surface forms, as shown below in \pref{fig:deletion-simple},
\pref{fig:deletion-except-1}, and \pref{fig:deletion-except-2}.

\begin{figure}[h]
\caption{Deletion in Yoruba}
\label{fig:deletion-simple}
\begin{tabular}{c|c|c|c}
    \textbf{Underlying} & \textbf{Surface} & \textbf{Underlying} & \textbf{Surface}\\
    \textipa{/ow\'o-k\'i-ow\'o/} & \textipa{[ow\'ok\'ow\'o]} & \textipa{/\textdyoghlig{}\'o-\`Ew\`u/} & \textipa{[\textdyoghlig{}\'Ew\`u]}\\
    \textipa{/OmO-k\'i-OmO/} & \textipa{[OmOk\'OmO]} & \textipa{/ra-\`Og\`Ed\`E/} & \textipa{[r\`Og\`Ed\`E]}\\
    \textipa{/se-ol\'u/} & \textipa{[sol\'u]} & \textipa{/n\'i-oko/} & \textipa{[l\'oko]}\\
    \textipa{/\textdyoghlig{}E-ed\'e/} & \textipa{[\textdyoghlig{}ed\'e]} & \textipa{/s\'i-O\textdyoghlig{}\`a/} & \textipa{[s\'O\textdyoghlig{}\`a]}\\
\end{tabular}
\end{figure}

% for handling [bimo], markedness constraint for preferring vowels to be
% approximately in the same place when both surrounding consonants share place?

However, [i] is always elided, regardless of whether it is the first or second
vowel in the sequence, as seen in \pref{fig:deletion-except-1}.

\begin{figure}[h]
\caption{Deletion of [i]}
\label{fig:deletion-except-1}
\begin{tabular}{c|c|c|c}
    \textbf{Underlying} & \textbf{Surface} & \textbf{Underlying} & \textbf{Surface}\\
    \textipa{/gb\'e-in\'O/} & \textipa{[gb\'en\'O]} & \textipa{/wo-il\`E/} & \textipa{[wol\`E]}\\
    \textipa{/gba-iS\'E/} & \textipa{[gbaS\'E]} & \textipa{/\textdyoghlig{}\'i-aSO/} & \textipa{[\textdyoghlig{}\'aSO]}\\
\end{tabular}
\end{figure}

However, when /u/ occurs at the end of the first morpheme it is always deleted,
unless followed by /i/ in which case /i/ is deleted, as seen in
\pref{fig:deletion-except-2}.

\begin{figure}[h]
\caption{Deletion of [u], except before [i]}
\label{fig:deletion-except-2}
\begin{tabular}{c|c|c|c}
    \textbf{Underlying} & \textbf{Surface} & \textbf{Underlying} & \textbf{Surface}\\
    \textipa{/\textdyoghlig{}u-igi/} & \textipa{[\textdyoghlig{}ugi]} & \textipa{/ru-epo/} & \textipa{[repo]}\\
    \textipa{/lu-il\`E/} & \textipa{[lul\`E]} & \textipa{/bu-omi/} & \textipa{[bomi]}\\
    \textipa{/b\'u-Oba/} & \textipa{[b\'Oba]}\\
\end{tabular}
\end{figure}

We can describe these three distinct groups of data straightforwardly. In
Yoruba, when two vowels become adjacent via concatenation of Yoruba morphemes,
the first vowel is deleted, unless the second vowel is less sonorous,
in which case it is deleted. This deletion occurs in order to ensure that
syllables always have an onset. This generalization can be analyzed from an
optimality-theoretic viewpoint. To begin, the phenomena in
\pref{fig:deletion-simple} can be naturally accounted for with a prohibition
against onsetless syllables, so one of the conjoined vowels is deleted to
prevent such a syllable from surfacing. Specifically, a `V`, with any number of
`C`s after it, is not a valid candidate for a syllable, because it lacks an
onset. We can codify this observation via the constraint \onset, which
precisely moderates against candidate surface forms with syllables which lack
an onset. This is defined in \pref{def:onset-and-maxc}.

% \textbf{make sure we're formulating the foot binarity effect correctly}

% \textbf{cite Yoruba Vowel Elisision: Minimality Effects}

\begin{figure}[h]
    \caption{Definitions for \onset{} and \maxc.}
    \label{def:onset-and-maxc}
    \begin{center}
        \onset\\
        Assign one violation mark for every syllable which lacks an onset.\\
        \vspace{1em}
        \maxc\\
        Assign one violation mark for every segment that is present in the
        underlying form but not in the surface form.
    \end{center}
\end{figure}

We can now establish a ranking between \onset{} and \maxc, which is the
classic faithfulness constraint responsible for preventing deletion and which
also has a definition in \pref{def:onset-and-maxc}. We will see later that
\maxc{} is used as a simplifying assumption, and that its effect here can
be accounted for better by a couple more complex constraints.

\begin{figure}[h]
    \caption{\onset{} >> \maxc}
    \label{tableau:onset-wins}
    \begin{tableau}{c|c}
        \inp{\ips{\textdyoghlig{}E-ed\'e}}       \const{\onset} \const{\maxc}
        \cand[\Optimal]{.\textdyoghlig{}e.d\'e.} \vio{}         \vio{*}
        \cand{.\textdyoghlig{}E.e.d\'e.}         \vio{*!}       \vio{}
    \end{tableau}
\end{figure}

This tableau proves that Yoruba prefers to delete one of the conjoined vowels
rather than create a syllable without an onset. However, a question immediately
arises: there is an unconsidered candidate, in this tableau, one where the
morpheme-initial vowel [e] was deleted instead of the morpheme-final vowel
\textipa{[E]}, to create the candidate surface form
\textipa{[\textdyoghlig{}E.d\'e]}. This also does not violate \onset, and has
an equal number of violations of \maxc, as shown in the tableau in
\pref{tableau:onset-maxc-tie}. Why didn't Yoruba choose this surface form
instead?

\begin{figure}[h]
    \caption{A tie between two candidates!}
    \label{tableau:onset-maxc-tie}
    \begin{tableau}{c|c}
        \inp{\ips{\textdyoghlig{}E-ed\'e}} \const{\onset} \const{\maxc}
        \cand{.\textdyoghlig{}e.d\'e.}     \vio{}         \vio{*}
        \cand{.\textdyoghlig{}E.d\'e}      \vio{}         \vio{*}
    \end{tableau}
\end{figure}

The answer lies in Yoruba's preference for preserving morpheme-initial vowels
versus morpheme-final vowels. Cross-linguistically, onsets are more important
than codas-- looking to the corpus of Optimality Theory constraints, for
instance, we see that \onset{} exists to regulate against syllables
\textit{without} onsets, but \nocoda{} exists to regulate against syllables
\textit{with} codas. Given this cross-linguistic asymmetry, it is natural to
conclude that Yoruba's treatment of vowels at morpheme boundaries (which are
also syllable boundaries) follows a similar pattern.

To break the tie in \pref{tableau:onset-maxc-tie}, we introduce a morphemic
faithfulness constraint against vowel deletion at the beginnings of morpheme
boundaries. We call this constraint \maxplusv. Though this means that we must
consult morphology in order to determine correct phonological outputs,
observing the phenomenon in \pref{fig:deletion-simple}, we see that it is
fundamentally a morphological one, so we posit that relying on morphology here
presents the most productive path forward. We also propose its sibling
constraint, \maxvplus, which prohibits vowels from being deleted in
morpheme-final positions. We define both of these constraints in
\pref{def:maxvplus-and-maxplusv}.

% \textbf{cite the paper from class as justification: onsets are generally more
% important than codas, onset vs nocoda}

\begin{figure}[h]
    \caption{Definitions for \maxplusv{} and \maxvplus.}
    \label{def:maxvplus-and-maxplusv}
    \begin{center}
        % Ident vowel-end-of-morpheme??
        \maxplusv\\
        Assign one violation mark for every vowel that is present at the
        beginning of a morpheme boundary in the UR and is not present in the
        SR.\\
        \vspace{1em}
        \maxvplus\\
        Assign one violation mark for every vowel that is present at the end of
        a morpheme boundary in the UR and is not present in the SR.
    \end{center}
\end{figure}

The tableau establishing their relative ranking is shown in
\pref{tableau:maxplusv-wins}.

\begin{figure}[h]
\caption{\maxplusv{} >> \maxvplus}
\label{tableau:maxplusv-wins}
\begin{tableau}{c|c}
    \inp{\ips{\textdyoghlig{}E-ed\'e}}    \const{\maxplusv} \const{\maxvplus}
    \cand[\Optimal]{\textdyoghlig{}ed\'e} \vio{}            \vio{*}
    \cand{\textdyoghlig{}Ed\'e}           \vio{*!}          \vio{}
\end{tableau}
\end{figure}

As \pref{tableau:maxplusv-wins} shows, Yoruba prefers morpheme final vowels to
be deleted over morpheme initial ones, because the deletion of the latter
constitutes a graver violation of \maxplusv. What this means is that we're not
\textit{really} dealing with \maxc{} as our primary faithfulness constraint, as
was implied by \pref{tableau:onset-wins}, but rather with our two faithfulness
constraints governing morpheme-initial versus morpheme-final vowel deletion. We
show an updated version of the tableau from \pref{tableau:onset-wins} in
\pref{tableau:onset-wins-v2} to account for this.

\begin{figure}[h]
    \caption{\onset{} >> \maxplusv}
    \label{tableau:onset-wins-v2}
    \begin{tableau}{c|c}
        \inp{\ips{\textdyoghlig{}E-ed\'e}}       \const{\onset} \const{\maxplusv}
        \cand[\Optimal]{.\textdyoghlig{}e.d\'e.} \vio{}         \vio{*}
        \cand{.\textdyoghlig{}E.e.d\'e.}         \vio{*!}       \vio{}
    \end{tableau}
\end{figure}

Our analysis so far accounts for the phenomena in \pref{fig:deletion-simple},
but not the phenomena in \pref{fig:deletion-except-1} or
\pref{fig:deletion-except-2}. We proceed, now, to analyze the first of these, in
which [i] is deleted regardless of whether it the morpheme initial or final
sound. One possible explanation as to why this is the case might be some
quality of the vowels themselves-- the frontness or backness of the vowels
appears to have no immediately discernible pattern with respect to deletion, so
we might try vowel height. Indeed, height appears to be correlated: [u] and
[i], which are both high vowels, are involved in the exceptional deletion
cases! We analyze both of these instances through the mechanism of sonority,
which offers an explanation for high vowels being involved in deletion-- they
are less sonorous. The sonority scale of vowels has been discussed in prior
work[\textbf{cite the quality-sensitive stress paper}, wikipedia article?].
This scale is shown in \pref{table:vowel-sonority}.

\begin{figure}[h]
    \caption{The vowel sonority scale (ordered by decreasing sonority)}
    \label{table:vowel-sonority}
    \begin{tabular}{c|c|c}
        \textbf{low} (\textsc{most} sonorous) & \textbf{mid} & \textbf{high} (\textsc{least} sonorous)\\
        /a/ & \textipa{/e E o O/} & /i u/\\
    \end{tabular}
\end{figure}

This helps inform why, for the data in \pref{fig:deletion-except-1}, [i] is
always elided, even though it appears morpheme-initially underlyingly.
Specifically, the language prefers to delete less sonorous vowels such as
[i] over more sonorous vowels. We can codify this intuition into a faithfulness
constraint, \ssp. This constraint is defined in \pref{def:ssp}.

% use smth like PEAK-PROMINENCE here instead?
\begin{figure}[h]
    \caption{Definition of \ssp}
    \label{def:ssp}
    \begin{center}
        \ssp\\
        For each vowel in the surface form, assign one additional violation
        mark for every increment of distance away from most sonorous level of
        the sonority scale.
    \end{center}
\end{figure}

We can now attempt to determine the ranking of this constraint that results in
the phenomenon seen in \pref{fig:deletion-except-1} while preserving the
phenomenon in \pref{fig:deletion-simple}.

\begin{figure}[h!]
    \caption{\ssp{} >> \maxplusv}
    \label{tableau:ssp-wins}
    \begin{tableau}{c|c}
        \inp{\ips{gb\'e-in\'O}}    \const{\ssp} \const{\maxplusv}
        \cand[\Optimal]{gb\'en\'O} \vio{**}     \vio{*}
        \cand{gb\'in\'O}           \vio{***!}   \vio{}
    \end{tableau}
\end{figure}

The tableau in \pref{tableau:ssp-wins} shows that the language prefers to
delete a morpheme-initial vowel over a more sonorous vowel (\textipa{[\'e]}
rather than \textipa{[\'i]}). This ranking completely account for the pairs
shown in \pref{fig:deletion-except-1}. We would additionally like to rank
\onset{} and \ssp{} against each other, but with the data we have, they cannot
be. For valid surface forms, only deletion ever occurs in order to satisfy
\onset{} or \ssp. Every time a vowel is deleted, violations of either of these
constraints cannot increase (for \onset, because there's one less syllable
nucleus which requires an onset, and for \ssp, because there's one less
potentially non-low vowel). As a result, we cannot attempt a ranking between
\onset{} and \ssp{} either.

We proceed to account, now, for the
final set of exceptions shown in \pref{fig:deletion-except-2}. We can actually
account for this phenomenon with a slight modification to our sonority scale,
ranking [u] above [i] within it.

\begin{figure}[h]
    \caption{The vowel sonority scale (ordered by decreasing sonority)}
    \label{table:vowel-sonority-v2}
    \begin{tabular}{c|c|c|c}
        \textbf{low} (\textsc{most} sonorous) & \textbf{mid} & \textbf{mid} &
            \textbf{high} (\textsc{least} sonorous)\\
        /a/ & \textipa{/e E o O/} & /u/ & /i/\\
    \end{tabular}
\end{figure}

This modification enables us to account for the phenomena in
\pref{fig:deletion-except-2} by ranking /i/ as lower in the sonority scale of
Yoruba. This elegantly deals with the issue of /i/ deletion at the beginning of
the second vowel being preferred over /u/ deletion at the end of the first
vowel, because, as discussed before, a violation of \ssp{} represents a
maximally severe violation of the language's phonology. This is explicated in
the tableau shown here.

% \begin{figure}[h]
%     \caption{Definition for \ssp{} (v2)}
%     \label{def:ssp-v2}
%     \begin{center}
%         \ssp\\
%         The Yoruba vowel sonority scale is defined as in 
%     \end{center}
% \end{figure}

\begin{figure}[h]
    \caption{\ssp >> \maxplusv (part 2)}
    \label{tableau:ssp-wins-v2}
    \begin{tableau}{c|c}
        \inp{\ips{lu-\`ilE}}    \const{\ssp} \const{\maxplusv}
        \cand[\Optimal]{lul\`E} \vio{***}    \vio{*}
        \cand{l\`il\`E}         \vio{****!}  \vio{}
    \end{tableau}
\end{figure}

As we can see here, the fact that morpheme-final [u] deletion is dispreferred
over morpheme-initial [i] deletion (which violates a fairly high-ranked
\maxplusv{} constraint) in the phenomena shown in \pref{fig:deletion-except-2}
is straightforwardly dealt with by taking the language-generic SSP ranking and
theorizing that Yoruba possesses a specialized variant of it, as many languages
do.

We now show a summary tableau for the input \textipa{/\textdyoghlig{}u-igi/}.

\begin{figure}[h!]
    \caption{Summary Tableau, attampt 1}
    \label{tableau:summary-one}
    \begin{tableau}{c:c|c|c}
        \inp{\ips{\textdyoghlig{}u-igi}}         \const{\onset} \const{\ssp}   \const{\maxplusv} \const{\maxvplus}
        \cand[\Optimal]{.\textdyoghlig{}u.gi.}   \vio{}         \vio{****!*}   \vio{*}           \vio{}
        \cand{.\textdyoghlig{}u.i.gi.}           \vio{*!}       \vio{*******}  \vio{}            \vio{}
        \cand{.\textdyoghlig{}i.gi.}             \vio{}         \vio{******!}  \vio{}            \vio{*}
        \cand{.\textdyoghlig{}a.gi.}             \vio{}         \vio{***}      \vio{*}           \vio{*}
    \end{tableau}
\end{figure}

We notice that candidate three is a winner threatening loser! It deletes both
of the non-sonorous vowels at the boundary of the two morphemes, and instead
epenthesizes an [a], which is a very sonorous vowel, incurring no violations of
\ssp. In order to prevent this winner-threatening loser, we introduce a
high-ranked \dep{} constraint. This is supported by our data, because we never
see epenthesis. We now show a simple ranking tableau that represents this argument.

\begin{figure}[h]
    \caption{\dep{} >> \ssp}
    \label{tableau:dep-wins}
    \begin{tableau}{c|c}
        \inp{\ips{\textdyoghlig{}u-igi}}       \const{\dep} \const{\ssp}
        \cand[\Optimal]{.\textdyoghlig{}u.gi.} \vio{}       \vio{*****}
        \cand{.\textdyoghlig{}a.gi.}           \vio{*!}     \vio{***}
    \end{tableau}
\end{figure}

We would additionally like to show a ranking for \dep{} over \onset, or
\onset{} over \dep, but unfortunately, this is impossible: we would need to
come up with a winning candidate that violates either \onset{} or \dep{},
neither of which is ever seen in the data.

Note that our winner-threatening loser could also have been created via a
featural change-- we can deal with this in a similar way, by introducing a
high-ranked \ident{} constraint. We show this below. For similar reasons to
that for \dep, \ident{} cannot be ranked against \onset. \dep{} and \ident{}
also cannot be ranked against each other, because we never see a surface
form that uses either.

\begin{figure}[h]
    \caption{\ident{} >> \ssp}
    \label{tableau:ident-wins}
    \begin{tableau}{c|c}
        \inp{\ips{\textdyoghlig{}u-igi}}       \const{\ident} \const{\ssp}
        \cand[\Optimal]{.\textdyoghlig{}u.gi.} \vio{}         \vio{*****}
        \cand{.\textdyoghlig{}a.gi.}           \vio{*!}       \vio{***}
    \end{tableau}
\end{figure}

And, finally, we can show our corrected summary tableau.

\begin{figure}[h]
    \caption{Final Summary Tableau}
    \label{tableau:summary-two}
    \resizebox{\textwidth}{!}{
    \begin{tableau}{c:c|c:c|c|c}
        \inp{\ips{\textdyoghlig{}u-igi}}       \const{\ident} \const{\dep} \const{\onset} \const{\ssp}   \const{\maxplusv} \const{\maxvplus}
        \cand[\Optimal]{.\textdyoghlig{}u.gi.} \vio{}         \vio{}       \vio{}         \vio{*****}    \vio{*}           \vio{}
        \cand{.\textdyoghlig{}u.i.gi.}         \vio{}         \vio{}       \vio{*!}       \vio{******!*} \vio{}            \vio{}
        \cand{.\textdyoghlig{}i.gi.}           \vio{}         \vio{}       \vio{}         \vio{******!}  \vio{}            \vio{*}
        \cand{.\textdyoghlig{}a.gi.}           \vio{*!}       \vio{*!}     \vio{}         \vio{***}      \vio{*}           \vio{*}
    \end{tableau}}
\end{figure}

Here, our previous winner-threatening-loser has a fatal violation of the
undominated \dep{} constraint, and so no longer threatens the winner!
We now have a complete ranking that accounts for all the data shown in
\pref{fig:deletion-simple}, \pref{fig:deletion-except-1}, and
\pref{fig:deletion-except-2}.

\begin{figure}[h]
\caption{Yoruba Surface Forms}
\label{fig:data}
\begin{tabular}{c c c c}
    \textbf{Word} & \textbf{Gloss} & \textbf{Word} & \textbf{Gloss}\\
    \textipa{ow\'o} & money & \textipa{k\'i} & any\\
    \textipa{OmO} & child & se & cook\\
    \textipa{ol\'u} & mushrooms & \textipa{\textdyoghlig{}E} & eat\\
    \textipa{ed\'e} & shrimp & \textipa{\textdyoghlig{}\'o} & burn\\
    \textipa{\`Ew\`u} & clothing & \textipa{Ra} & buy\\
    \textipa{\`Og\`Ed\`E} & banana & \textipa{n\'i} & at\\
    \textipa{s\'i} & to & \textipa{O\textdyoghlig{}\`a} & market\\
    \textipa{ow\'ok\'ow\'o} & any money at all | bad money & \textipa{\textdyoghlig{}ed\'e} & eat shrimp\\
    \textipa{OmOk\'OmO} & any child at all | bad child & \textipa{sol\'u} & cook mushrooms\\
    \textipa{\textdyoghlig{}\'Ew\`u} & burn clothing & \textipa{R\`Og\`Ed\`E} & buy bananas\\
    \textipa{l\'oko} & at the farm & \textipa{s\'O\textdyoghlig{}\`a} & to the market\\
    \textipa{gb\'e} & lift & \textipa{in\'O} & lamp\\
    \textipa{gba} & take & \textipa{iS\'E} & job\\
    \textipa{wo} & look & \textipa{il\`E} & ground\\
    \textipa{\textdyoghlig{}\'i} & steal & \textipa{aSO} & clothes\\
    \textipa{gb\'en\'O} & lift the lamp & \textipa{gbaS\'E} & take a job\\
    \textipa{wol\`E} & look at the ground & \textipa{\textdyoghlig{}\'aSO} & steal clothes\\
    \textipa{\textdyoghlig{}u} & throw & \textipa{igi} & stick\\
    \textipa{lu} & hit & \textipa{il\`E} & ground\\
    \textipa{bu} & take & \textipa{omi} & water\\
    \textipa{Ru} & carry & \textipa{epo} & oil\\
    \textipa{b\'u} & abuse & \textipa{Oba} & king\\
    \textipa{\textdyoghlig{}ugi} & throw a stick & \textipa{lul\`E} & hit the ground\\
    \textipa{bomi} & take water & \textipa{Repo} & carry oil\\
    \textipa{b\'Oba} & abuse the king & \textipa{b\'i} & have\\
    \textipa{b\'imO} & have a baby\\
\end{tabular}
\end{figure}

\end{document}

### 2023-01-29 proposal.tex
Add initial part of proposal for capstone

\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}

\title{A computational analysis of vowel elision in Yoruba collocations}
\author{Hemant Gouni}

\begin{document}

\maketitle

In Yoruba, deletion occurs at morpheme boundaries. Specifically, when two
vowels become adjacent via concatenation of Yoruba morphemes, the first vowel
is generally deleted. However, [i] is always elided, regardless of whether it
is the first or second vowel in the sequence. Furthermore, however, when /u/
occurs at the end of the first morpheme it is always deleted, unless followed
by /i/ in which case /i/ is deleted. This is a well-known phenomenon to which
I provided an Optimality-Theoretic analysis in my previous paper.

I would like to extend this work in a further paper, not necessarily by
refining the theoretical basis of the analysis, but by offering a computational
interpretation of it. Specifically, I would like to create a program which,
when given a Yoruba underlying representation, creates the appropriate surface
form using the constraints and rankings derived in my analysis. This would be
an interesting project to explore because it may offer some insight into the
mechanical aspects of Optimality Theory which are not often addressed in the
literature. For instance, Optimality Theory relies on \textsc{Gen}, a mechanism
that produces all posssible surface forms from a given underlying one. These
forms are then evaluated in parallel against a given constraint ranking, with
the best ranked one being chosen as the winner. However, clearly, \textsc{Gen}
cannot truly produce all possible candidate output forms, because this would
take infinite time. Determining a heuristic that approximates \textsc{Gen}'s
functionality efficiently would be one such interesting aspect of this work.
The analysis from my previous paper would merely be used as a running example
to motivate and validate the results of the computational analysis, because I
am already familiar with it.

To summarize, the question I want to pursue is this: "What additional
considerations must be added to a standard Optimality-theoretic analysis in
order to make it amenable to automation?" I plan to start this work by
surveying any previous research on computer-aided phonological analysis,
particularly in the context of Optimality Theory. I will then attempt to write
my own computational analysis, perhaps using a language especially well-suited
to constraint-based programming like Prolog or Haskell, if I can find a way to
translate the necessary constructs (between my analysis and the language)
appropriately. If I succeed at doing this, I will then embark on the more

\end{document}

### 2023-02-25 paper.tex
Add basic intro, some discussion on approach

@@
 \usepackage[margin=1in]{geometry}
+
+\usepackage{biblatex}
+\addbibresource{references.bib}
 
@@
 \maketitle
+
+\section{Analysis}
 
@@
 syllables always have an onset. This generalization can be analyzed from an
-optimality-theoretic viewpoint. To begin, the phenomena in
+Optimality-theoretic viewpoint. To begin, the phenomena in
 \pref{fig:deletion-simple} can be naturally accounted for with a prohibition
@@
 versus morpheme-final vowels. Cross-linguistically, onsets are more important
-than codas-- looking to the corpus of Optimality Theory constraints, for
+than codas--- looking to the corpus of Optimality Theory constraints, for
 instance, we see that \onset{} exists to regulate against syllables
@@
 sound. One possible explanation as to why this is the case might be some
-quality of the vowels themselves-- the frontness or backness of the vowels
+quality of the vowels themselves--- the frontness or backness of the vowels
 appears to have no immediately discernible pattern with respect to deletion, so
@@
 cases! We analyze both of these instances through the mechanism of sonority,
-which offers an explanation for high vowels being involved in deletion-- they
+which offers an explanation for high vowels being involved in deletion--- they
 are less sonorous. The sonority scale of vowels has been discussed in prior
@@
 Note that our winner-threatening loser could also have been created via a
-featural change-- we can deal with this in a similar way, by introducing a
+featural change--- we can deal with this in a similar way, by introducing a
 high-ranked \ident{} constraint. We show this below. For similar reasons to
@@
 
+\pagebreak
+
 \begin{figure}[h]
@@
 
+\pagebreak
+
+\section{Computing Ideal Surface Forms}
+
+\subsection{Introduction and Background}
+
+We now turn to assigning a computational interpretation to our analysis.
+Creating an algorithm to compute ideal outputs from Optimality-theoretic
+constraints has been explored in prior work, though not extensively. Kartunnen
+\cite{karttunen2006finite} explores a computational implementation of the
+Optimality-theoretic analysis of Finnish stress using the finite state machine
+description language XFST (\cite{karttunen2001applications} gives a broad
+overview of finite state transducers as applied to morphological analysis and
+parsing). Interestingly, such an implementation leads to insights about the
+soundness of the preexisting analyses of Finnish stress--- namely, that they
+are not. Key to this analysis is that it is an \textit{approximation}: indeed,
+the paper itself notes that ``it has been known for a long time [\ldots] that
+\textsc{OT} is not a finite-state system'', and that ``\textsc{OT} is
+fundamentally more complex than [rule-based phonological systems]''.
+
+Idsardi \cite{idsardi2006simple} explores this computational aspect, providing
+a proof that generating an appropriate candidate set which can be evaluated
+against some constraint ranking is \textit{NP-Hard}. This is a term used in
+computability theory to describe a problem to which any number of other hard
+problems can be \textit{reduced}. In essence, if there exists an efficient way
+to solve this problem, a number of other incredibly difficult problems could
+also be easily solved. Knowing a problem is NP-Hard is strong evidence that it
+cannot be solved efficiently.
+
+Heinz, Kobele, and Riggle \cite{heinz2009evaluating} contest this result,
+showing that a different representation of \textsc{OT} grammars as a single
+finite state transducer (instead of a sequence of finite state constriants)
+averts the computational intractability issue demonstrated in
+\cite{idsardi2006simple}. Additionally, they outline two subclasses of the
+\textit{universal problem} of attempting to determine a valid constraint
+ranking and underlying representations given only surface forms. Namely,
+the \textit{simple problem} asks for optimal surface forms given an underlying
+form and fixed constraint set \textsc{Con} and ranking \textsc{R}. The
+\textsc{quasi-universal problem} asks for an optimal surface form given
+a fixed constraint set with any possible ranking.
+
+Encoding our analysis as an algorithm has several advantages, namely, that we
+make explicit the procedure for calculating surface forms from underlying ones,
+forcing us to make obvious any implicit assumptions made in the analysis as we
+step through it. Secondly, and perhaps more interestingly, this enables us to
+test our analysis quickly and thoroughly given new data, ensuring that it holds
+and giving us opportunities to improve it otherwise.
+
+\subsection{Approach}
+
+We base our procedure, crucially, on bounded depth-first search. We first
+describe the basic algorithm, before proceeding onwards to an implementation.
+Viewed as a single function which takes an input and returns an output, our
+analysis takes as input an underlying form and a set of ranked constraints,
+and searches for the correct output form. The search for the correct output
+is of course the crucial portion here, and is the part on which most of the
+discussion will be spent.
+
+Our basic approach employs \textit{bounded depth-first search}. It is best to
+illustrate this technique on a concrete example. Take, for instance, the
+surface form \textipa{[sol\'u]} and its underlying representation
+\textipa{/se-ol\'u/}. We will demonstrate our technique on this mapping.
+
+Our first order of business is to generate a reasonably large set of
+permutations of \textipa{[sol\'u]}, one of which is the correct output form. We
+are not yet attempting to identify which is the correct output form, only
+generate it so it may be picked later. In effect, we are attempting to
+efficiently emulate Optimality Theory's \textsc{Gen}. To do this, we intoduce
+the following core transformations; we model each of these transformations as
+functions of their own:
+
+\begin{enumerate}
+    \item \textbf{Delete}: Delete a segment
+    \item \textbf{Epenthesize}: Insert a segment
+    \item \textbf{Change}: Map a segment to another one
+    \item \textbf{Metathesize}: Swap the positions of two segments
+\end{enumerate}
+
+\textbf{Delete} takes as input a form, such as \textipa{[sol\'u]}, and
+randomly deletes one of the characters. For instance, it may delete
+\textipa{[o]} to produce \textipa{[sl\'u]}, though again, this says little
+about whether such a form would surface. It is more likely to delete inside the
+word than at the edges of the word, following in line with broad phonological
+generalizations about the markedness of deleting morpheme-initial or
+morpheme-final segments (citation about this here?).
+
+\textbf{Epenthesize} inserts a segment into its input and returns the new form.
+For example, if given \textipa{[sol\'u]} as input, epenthesis may return
+\textipa{[sool\'u]}. Again, it is more likely to epenthesize inside the form
+than at its edges.
+
+\textbf{Change}, to some extent, is functionally a composition of the previous
+two operations, replacing one segment with another one. However, this does not
+reflect how such transitions are modeled in the literature, using the
+categorization of segments into their features. To accommodate this without
+undertaking the modeling of autosegmental phonology
+\cite{goldsmith1990autosegmental} in a computationally tractable way, we
+manually partition the set of segments into phonologically similar (not
+necessarily disjoint) sets, and attempt to map largely within these sets.
+Illustrating concretely, we might classify all vowels in the language into one
+set, and all the sonorants into another, so our two sets would look something
+like this:
+
+\subsection{Implementation}
+
+\subsection{Analysis on corpus data}
+
+\section{Concluding}
+
+\printbibliography
+
 \end{document}

### 2023-02-26 paper.tex
Add newest changes

@@
 
-\section{Computing Ideal Surface Forms}
+\section{Computing Optimal Surface Forms}
 
@@
 
+\subsubsection{Core Operations for Generating Candidates}
+
 Our first order of business is to generate a reasonably large set of
@@
 Illustrating concretely, we might classify all vowels in the language into one
-set, and all the sonorants into another, so our two sets would look something
-like this:
+set, and all sonorants into another:
+
+\begin{enumerate}
+    \item \textipa{a o O i E u}
+    \item \textipa{l m n w j}
+\end{enumerate}
+
+The probability of a segment changing internally to either of these sets is
+high, but the probability of a segment changing across one or the other is much
+lower. We note that it may be necessary to adjust the weighting of these
+probabilities based on the language, or on cross-linguistic data more generally.
+
+\textbf{Metathesize}, finally, can similarly be viewed as a composition of the
+delete and epenthesize, or as a special case of the change operation. It
+differs, of course, in its ability to make targeted transpositions which are
+thought, based on cross-linguistic data, to be atomic (meaning indivisible, or
+inseparable into components) operations of their own. In particular, it's
+significantly more likely to swap two adjacent segments, again deploying
+cross-linguistic observations.
+
+\subsubsection{Generating Candidates}
+
+With these operations in hand, we're now ready to address the interesting
+portion: generating candidates and identifying the ideal one. First up, a short
+description of how candidates are generated. We start with the underlying
+representation. We then use each of \texttt{delete}, \texttt{epenthesize},
+\texttt{change}, and \texttt{metathesize} on the underlying form, 4 times each.
+This produces an initial candidate set of 16 candidates. We repeat the process
+again on each entry, producing 256 entries. We repeat two more times, yielding
+65536 total candidates. We're now ready to evaluate this candidate set against
+our ranked constraints, but before we do so, it is appropriate to discuss how,
+precisely, we represent constraints in a computable way.
+
+\subsubsection{Encoding Constraints}
+
+We represent constraints as functions. Specifically, these functions take as
+input a candidate surface form, compare it against the underlying
+representation, and return a count of violations. The precise logic used to
+compare a candidate surface form against the underlying representation will
+depend on the nature of each constraint. For example, for the constraint
+\textsc{Dep}, the function would compare the underlying form to the candidate
+surface form and determine if the latter has any segments which are not present
+in the former. Of course, it must also determine whether the seemingly
+epenthetic segment is the result of \texttt{change} or \texttt{metathesize},
+and if so, avoid outputting a violation mark for that segment. This could be
+determined by comparing segments around the errant one for equality. If
+corresponding segements are equal at corresponding indices of the underlying
+representation and the candidate surface form, modulo the errant segment, we
+can be reasonably sure that \texttt{change} has taken place, otherwise, if
+they are not, we can be reasonably sure that \texttt{metathesize} is the
+culprit.
+
+\subsubsection{Evaluating Candidates}
+
+We are now in the appropriate setting to identify the correct output form.
+In order to do this, we must evaluate our entire set of constraints against
+all of the 65536 candidate surface forms we generated previously. This will
+generate a sequence of constraint violations, one for each candidate surface
+form.
 

### 2023-02-26 paper.tex
Description of algorithm sorta complete

@@
 
-We are now in the appropriate setting to identify the correct output form.
-In order to do this, we must evaluate our entire set of constraints against
-all of the 65536 candidate surface forms we generated previously. This will
-generate a sequence of constraint violations, one for each candidate surface
-form.
+We are now in the appropriate setting to identify the correct output form. In
+order to do this, we must evaluate our entire set of constraints against all of
+the 65536 candidate surface forms we generated previously. This will generate a
+sequence of constraint violations, one for each candidate surface form. We
+search through each sequence in parallel, looking at the number of violations
+for each constraint at the same time, starting from the highest-ranked. We
+determine the minimal number of violations for the current constraint, and
+eliminate all candidates with violation counts that exceed this threshold. We
+repeat until we have a single candidate left, and return this candidate.
+
+TODO: describe the algorithm succinctly using an algorithm figure?
 
@@
 
+TODO: Figure out how best to implement the algorithm described above.
+
+TODO: Show code examples of the critical parts, one for each of the
+subsections above!
+
 \subsection{Analysis on corpus data}
+
+In this section we will evaluate all of our collected Yoruba underlying
+representations against the expected surface forms.
+
+TODO: Requires implementation to be reasonably far along first!
 

### 2023-03-03 paper.tex
Add source, fix formatting

@@
 \usepackage[margin=1in]{geometry}
+
+\usepackage{setspace}
+% \doublespacing
 
@@
 are less sonorous. The sonority scale of vowels has been discussed in prior
-work[\textbf{cite the quality-sensitive stress paper}, wikipedia article?].
-This scale is shown in \pref{table:vowel-sonority}.
+work \cite{kenstowicz1997quality}. This scale is shown in
+\pref{table:vowel-sonority}.
 
@@
 form and fixed constraint set \textsc{Con} and ranking \textsc{R}. The
-\textsc{quasi-universal problem} asks for an optimal surface form given
+\textit{quasi-universal problem} asks for an optimal surface form given
 a fixed constraint set with any possible ranking.

### 2023-03-05 yoruba-ot/src/main.rs
Add initial parts of implementation for ot

@@ -1,0 +1,83 @@
+use rand::{rngs::StdRng, Rng, SeedableRng};
+use unicode_segmentation::UnicodeSegmentation;
+
+#[derive(Clone, Debug)]
+struct Candidate {
+    form: String,
+    rng: StdRng,
+}
+
+impl Candidate {
+    fn len(&self) -> usize {
+        self.form.graphemes(true).collect::<Vec<&str>>().len()
+    }
+
+    fn delete(&self) -> Self {
+        let mut self_copy = self.clone();
+
+        let index = if self_copy.len() > 0 {
+            self_copy.rng.gen_range(0..self_copy.len())
+        } else {
+            0
+        };
+
+        self_copy.form = self_copy
+            .form
+            .grapheme_indices(true)
+            .filter(|(char_index, _)| *char_index != index)
+            .map(|(_, str)| str)
+            .collect::<String>();
+
+        self_copy
+    }
+
+    fn epenthesize(&self) -> Self {
+        let mut self_copy = self.clone();
+        let index = self_copy.rng.gen_range(0..self_copy.form.len());
+        self_copy.form.remove(index);
+        self_copy
+    }
+}
+
+fn main() {
+    let rng = StdRng::seed_from_u64(7777777);
+    let cand = Candidate {
+        form: "test".to_string(),
+        rng,
+    };
+    println!("{:?}", cand.delete().delete().delete());
+}
+
+#[cfg(test)]
+mod test {
+    use super::*;
+    use once_cell::sync::Lazy;
+
+    static RNG: Lazy<StdRng> = Lazy::new(|| StdRng::seed_from_u64(7777777));
+
+    #[test]
+    fn test_delete_1() {
+        let cand = Candidate {
+            form: "test".to_string(),
+            rng: RNG.clone(),
+        }
+        .delete();
+
+        assert_eq!(cand.form, "tst");
+    }
+
+    #[test]
+    fn test_delete_2() {
+        let cand = Candidate {
+            form: "test".to_string(),
+            rng: RNG.clone(),
+        }
+        .delete()
+        .delete()
+        .delete()
+        .delete()
+        .delete();
+
+        assert_eq!(cand.form, "");
+    }
+}

### 2023-03-19 yoruba-ot/src/main.rs
Add recent changes, brush up deleting

@@ -1,22 +1,25 @@
 use rand::{rngs::StdRng, Rng, SeedableRng};
 use unicode_segmentation::UnicodeSegmentation;
 
-#[derive(Clone, Debug)]
+#[derive(Debug, Clone)]
 struct Candidate {
     form: String,
     rng: StdRng,
 }
 
 impl Candidate {
-    fn len(&self) -> usize {
-        self.form.graphemes(true).collect::<Vec<&str>>().len()
+    fn indices(&self) -> Vec<usize> {
+        self.form
+            .grapheme_indices(true)
+            .map(|(index, _)| index)
+            .collect::<Vec<usize>>()
     }
 
-    fn delete(&self) -> Self {
+    fn delete(&mut self) -> Self {
         let mut self_copy = self.clone();
 
-        let index = if self_copy.len() > 0 {
-            self_copy.rng.gen_range(0..self_copy.len())
+        let index = if self_copy.indices().len() > 0 {
+            self_copy.indices()[self.rng.gen_range(0..self_copy.indices().len())]
         } else {
             0
         };
@@ -30,36 +33,33 @@ impl Candidate {
 
         self_copy
     }
-
-    fn epenthesize(&self) -> Self {
-        let mut self_copy = self.clone();
-        let index = self_copy.rng.gen_range(0..self_copy.form.len());
-        self_copy.form.remove(index);
-        self_copy
-    }
 }
 
+trait Constraint {
+    fn evaluate(candidate: Candidate) -> usize;
+}
+
+// impl Constraint {
+
+// }
+
 fn main() {
-    let rng = StdRng::seed_from_u64(7777777);
-    let cand = Candidate {
-        form: "test".to_string(),
-        rng,
+    let mut cand = Candidate {
+        form: "owókíowó".to_string(),
+        rng: StdRng::seed_from_u64(7777777),
     };
-    println!("{:?}", cand.delete().delete().delete());
+    println!("{:#?}", cand.delete().delete().delete().delete());
 }
 
 #[cfg(test)]
 mod test {
     use super::*;
-    use once_cell::sync::Lazy;
-
-    static RNG: Lazy<StdRng> = Lazy::new(|| StdRng::seed_from_u64(7777777));
 
     #[test]
     fn test_delete_1() {
         let cand = Candidate {
             form: "test".to_string(),
-            rng: RNG.clone(),
+            rng: StdRng::seed_from_u64(7777777),
         }
         .delete();
 
@@ -70,8 +70,40 @@ mod test {
     fn test_delete_2() {
[…]

### 2023-04-12 yoruba-ot/src/main.rs
Begin refactor to handle syllables

@@ -1,6 +1,34 @@
+#![allow(dead_code)]
+
 use rand::{rngs::StdRng, Rng, SeedableRng};
 use similar::{DiffOp, TextDiff};
 use unicode_segmentation::UnicodeSegmentation;
+
+// string -> syllabified candidate -> random deletions (all winners generated via deletions) ->
+// eval against constraints
+
+const VOWELS: [&'static str; 7] = ["o", "ɛ", "ɔ", "i", "u", "a", "e"];
+
+#[derive(Debug, Clone)]
+struct SyllabifiedCandidate {
+    form: Vec<Segment>,
+    rng: StdRng,
+}
+
+#[derive(Debug, Clone)]
+struct Segment {
+    char: String,
+    char_type: SegmentType,
+}
+
+#[derive(Debug, Clone)]
+enum SegmentType {
+    Onset,
+    Vowel,
+    Coda,
+}
+
+// fn syllabify()
 
 #[derive(Debug, Clone)]
 struct Candidate {
@@ -16,6 +44,7 @@ impl Candidate {
             .collect::<Vec<usize>>()
     }
 
+    // &mut because we want to change the rng when we delete
     fn delete(&mut self) -> Self {
         let mut self_copy = self.clone();
 
@@ -71,6 +100,15 @@ impl Constraint for Dep {
             .count()
     }
 }
+
+struct Onset(Candidate);
+
+// impl Constraint for Onset {
+//     fn evaluate(self, surface: Candidate) -> usize {
+//         let diff = TextDiff::from_graphemes(&self.0.form, &surface.form);
+
+//     }
+// }
 
 fn main() {
     let cand1 = Candidate {

### 2023-04-13 yoruba-ot/src/main.rs
Rough implementation of syllabification, ONSET

@@ -1,8 +1,12 @@
 #![allow(dead_code)]
 
+mod utils;
+
 use rand::{rngs::StdRng, Rng, SeedableRng};
 use similar::{DiffOp, TextDiff};
 use unicode_segmentation::UnicodeSegmentation;
+
+use utils::PushRet;
 
 // string -> syllabified candidate -> random deletions (all winners generated via deletions) ->
 // eval against constraints
@@ -21,59 +25,153 @@ struct Segment {
     char_type: SegmentType,
 }
 
-#[derive(Debug, Clone)]
+#[derive(Debug, Clone, Eq, PartialEq)]
 enum SegmentType {
     Onset,
-    Vowel,
+    Nucleus,
     Coda,
+    None,
 }
 
-// fn syllabify()
-
-#[derive(Debug, Clone)]
-struct Candidate {
-    form: String,
-    rng: StdRng,
-}
-
-impl Candidate {
-    fn indices(&self) -> Vec<usize> {
-        self.form
-            .grapheme_indices(true)
-            .map(|(index, _)| index)
-            .collect::<Vec<usize>>()
-    }
-
-    // &mut because we want to change the rng when we delete
-    fn delete(&mut self) -> Self {
-        let mut self_copy = self.clone();
-
-        let index = if self_copy.indices().len() > 0 {
-            self_copy.indices()[self.rng.gen_range(0..self_copy.indices().len())]
-        } else {
-            0
-        };
-
-        self_copy.form = self_copy
-            .form
-            .grapheme_indices(true)
-            .filter(|(char_index, _)| *char_index != index)
-            .map(|(_, str)| str)
-            .collect::<String>();
-
-        self_copy
+impl SyllabifiedCandidate {
+    fn delete(mut self) -> Self {
+        if self.form.len() > 0 {
+            self.form.remove(self.rng.gen_range(0..self.form.len()));
+        }
+        self
     }
 }
+
+impl From<&str> for SyllabifiedCandidate {
+    fn from(str: &str) -> SyllabifiedCandidate {
+        SyllabifiedCandidate {
+            form: syllabify(
+                str.graphemes(true)
+                    .map(|grapheme| Segment {
+                        char: grapheme.to_owned(),
+                        char_type: SegmentType::None,
+                    })
+                    .collect(),
+            ),
+            rng: StdRng::seed_from_u64(7777777),
+        }
+    }
+}
+
+impl From<SyllabifiedCandidate> for String {
[…]

### 2023-04-13 yoruba-ot/src/utils.rs

@@ -1,0 +1,10 @@
+pub trait PushRet<T> {
+    fn push_ret(&mut self, t: T) -> &mut Vec<T>;
+}
+
+impl<T> PushRet<T> for Vec<T> {
+    fn push_ret(&mut self, t: T) -> &mut Vec<T> {
+        self.push(t);
+        self
+    }
+}

### 2023-04-13 yoruba-ot/src/main.rs
Address Clippy warnings

@@ -11,7 +11,7 @@ use utils::PushRet;
 // string -> syllabified candidate -> random deletions (all winners generated via deletions) ->
 // eval against constraints
 
-const VOWELS: [&'static str; 7] = ["o", "ɛ", "ɔ", "i", "u", "a", "e"];
+const VOWELS: [&str; 7] = ["o", "ɛ", "ɔ", "i", "u", "a", "e"];
 
 #[derive(Debug, Clone)]
 struct SyllabifiedCandidate {
@@ -35,7 +35,7 @@ enum SegmentType {
 
 impl SyllabifiedCandidate {
     fn delete(mut self) -> Self {
-        if self.form.len() > 0 {
+        if !self.form.is_empty() {
             self.form.remove(self.rng.gen_range(0..self.form.len()));
         }
         self
@@ -125,41 +125,6 @@ fn syllabify(candidate: Vec<Segment>) -> Vec<Segment> {
     mark_codas(mark_onsets(mark_vowels(candidate)))
 }
 
-// #[derive(Debug, Clone)]
-// struct Candidate {
-//     form: String,
-//     rng: StdRng,
-// }
-
-// impl Candidate {
-//     fn indices(&self) -> Vec<usize> {
-//         self.form
-//             .grapheme_indices(true)
-//             .map(|(index, _)| index)
-//             .collect::<Vec<usize>>()
-//     }
-
-//     // &mut because we want to change the rng when we delete
-//     fn delete(&mut self) -> Self {
-//         let mut self_copy = self.clone();
-
-//         let index = if self_copy.indices().len() > 0 {
-//             self_copy.indices()[self.rng.gen_range(0..self_copy.indices().len())]
-//         } else {
-//             0
-//         };
-
-//         self_copy.form = self_copy
-//             .form
-//             .grapheme_indices(true)
-//             .filter(|(char_index, _)| *char_index != index)
-//             .map(|(_, str)| str)
-//             .collect::<String>();
-
-//         self_copy
-//     }
-// }
-
 trait Constraint {
     fn evaluate(self, surface: SyllabifiedCandidate) -> usize;
 }
@@ -175,10 +140,7 @@ impl Constraint for Ident {
 
         diff.ops()
             .iter()
-            .filter(|op| match op {
-                DiffOp::Replace { .. } => true,
-                _ => false,
-            })
+            .filter(|op| matches!(op, DiffOp::Replace { .. }))
             .count()
     }
 }
[…]

### 2023-04-13 yoruba-ot/src/main.rs
Add SSP

@@ -22,15 +22,22 @@ struct SyllabifiedCandidate {
 #[derive(Debug, Clone)]
 struct Segment {
     char: String,
-    char_type: SegmentType,
+    syllable_index: SyllableIndex,
+    seg_type: SegmentType,
 }
 
 #[derive(Debug, Clone, Eq, PartialEq)]
-enum SegmentType {
+enum SyllableIndex {
     Onset,
     Nucleus,
     Coda,
     None,
+}
+
+#[derive(Debug, Clone, Eq, PartialEq)]
+enum SegmentType {
+    Vowel,
+    Consonant,
 }
 
 impl SyllabifiedCandidate {
@@ -42,6 +49,14 @@ impl SyllabifiedCandidate {
     }
 }
 
+fn get_seg_type(grapheme: &str) -> SegmentType {
+    if VOWELS.contains(&grapheme) {
+        SegmentType::Vowel
+    } else {
+        SegmentType::Consonant
+    }
+}
+
 impl From<&str> for SyllabifiedCandidate {
     fn from(str: &str) -> SyllabifiedCandidate {
         SyllabifiedCandidate {
@@ -49,7 +64,8 @@ impl From<&str> for SyllabifiedCandidate {
                 str.graphemes(true)
                     .map(|grapheme| Segment {
                         char: grapheme.to_owned(),
-                        char_type: SegmentType::None,
+                        syllable_index: SyllableIndex::None,
+                        seg_type: get_seg_type(grapheme),
                     })
                     .collect(),
             ),
@@ -65,10 +81,11 @@ impl From<SyllabifiedCandidate> for String {
 }
 
 impl Segment {
-    fn morph_type(&self, seg_type: SegmentType) -> Segment {
+    fn morph_type(&self, seg_type: SyllableIndex) -> Segment {
         Segment {
             char: self.char.clone(),
-            char_type: seg_type,
+            syllable_index: seg_type,
+            seg_type: self.seg_type.clone(),
         }
     }
 }
@@ -83,7 +100,7 @@ fn mark_vowels(candidate: Vec<Segment>) -> Vec<Segment> {
         .iter()
         .map(|segment| {
             if VOWELS.contains(&segment.char.as_str()) {
-                segment.morph_type(SegmentType::Nucleus)
+                segment.morph_type(SyllableIndex::Nucleus)
             } else {
                 segment.clone()
             }
@@ -92,15 +109,16 @@ fn mark_vowels(candidate: Vec<Segment>) -> Vec<Segment> {
 }
 
[…]

### 2023-04-13 yoruba-ot/src/main.rs
Rename SonoritySequencingPrinciple to SonSeqPr

@@ -199,9 +199,9 @@ impl Constraint for Onset {
     }
 }
 
-struct SonoritySequencingPrinciple;
+struct SonSeqPr;
 
-impl Constraint for SonoritySequencingPrinciple {
+impl Constraint for SonSeqPr {
     fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
         surface
             .form
@@ -222,7 +222,7 @@ impl Constraint for SonoritySequencingPrinciple {
 
 fn main() {
     let syllabified_candidate: SyllabifiedCandidate = "owókíowó".into();
-    dbg!(SonoritySequencingPrinciple.evaluate(syllabified_candidate));
+    dbg!(SonSeqPr.evaluate(syllabified_candidate));
 }
 
 #[cfg(test)]
@@ -288,7 +288,7 @@ mod test {
     fn test_ssp_1() {
         let syllabified_candidate: SyllabifiedCandidate = "owókíowó".into();
         assert_eq!(
-            SonoritySequencingPrinciple.evaluate(syllabified_candidate),
+            SonSeqPr.evaluate(syllabified_candidate),
             7
         );
     }
@@ -297,7 +297,7 @@ mod test {
     fn test_ssp_2() {
         let syllabified_candidate: SyllabifiedCandidate = "".into();
         assert_eq!(
-            SonoritySequencingPrinciple.evaluate(syllabified_candidate),
+            SonSeqPr.evaluate(syllabified_candidate),
             0
         );
     }

### 2023-04-13 yoruba-ot/src/main.rs
Handle affricates

@@ -59,16 +59,26 @@ fn get_seg_type(grapheme: &str) -> SegmentType {
 
 impl From<&str> for SyllabifiedCandidate {
     fn from(str: &str) -> SyllabifiedCandidate {
+        let mut graphemes: Vec<Segment> = str
+            .graphemes(true)
+            .map(|grapheme| Segment {
+                char: grapheme.to_owned(),
+                syllable_index: SyllableIndex::None,
+                seg_type: get_seg_type(grapheme),
+            })
+            .collect();
+
+        // grouping affricates together as non-separate segments
+        for idx in 0..(graphemes.len() - 1) {
+            if graphemes[idx].char == "d\u{361}" && graphemes[idx + 1].char == "ʒ" {
+                let char_to_concat = graphemes[idx + 1].char.clone();
+                graphemes[idx].char.push_str(&char_to_concat);
+                graphemes.remove(idx + 1);
+            }
+        }
+
         SyllabifiedCandidate {
-            form: syllabify(
-                str.graphemes(true)
-                    .map(|grapheme| Segment {
-                        char: grapheme.to_owned(),
-                        syllable_index: SyllableIndex::None,
-                        seg_type: get_seg_type(grapheme),
-                    })
-                    .collect(),
-            ),
+            form: syllabify(graphemes),
             rng: StdRng::seed_from_u64(7777777),
         }
     }
@@ -210,7 +220,7 @@ impl Constraint for SonSeqPr {
             //
             // .next().unwrap() should never panic here bc that's only possible if the initial
             // candidate input string is empty, and if that's true, then the iterator will be empty
-            .map(|seg| match seg.char.chars().next().unwrap() {
+            .map(|seg| match dbg!(seg.char.chars().next().unwrap()) {
                 'e' | 'ɛ' | 'o' | 'ɔ' => 1,
                 'u' => 2,
                 'i' => 3,
@@ -221,7 +231,7 @@ impl Constraint for SonSeqPr {
 }
 
 fn main() {
-    let syllabified_candidate: SyllabifiedCandidate = "owókíowó".into();
+    let syllabified_candidate: SyllabifiedCandidate = dbg!("d͡ʒuigi".into());
     dbg!(SonSeqPr.evaluate(syllabified_candidate));
 }
 
@@ -287,18 +297,12 @@ mod test {
     #[test]
     fn test_ssp_1() {
[…]

### 2023-04-13 yoruba-ot/src/main.rs
Bugfix

@@ -69,11 +69,14 @@ impl From<&str> for SyllabifiedCandidate {
             .collect();
 
         // grouping affricates together as non-separate segments
-        for idx in 0..(graphemes.len() - 1) {
-            if graphemes[idx].char == "d\u{361}" && graphemes[idx + 1].char == "ʒ" {
-                let char_to_concat = graphemes[idx + 1].char.clone();
-                graphemes[idx].char.push_str(&char_to_concat);
-                graphemes.remove(idx + 1);
+        // can we do this more cleanly?
+        if !graphemes.is_empty() {
+            for idx in 0..graphemes.len() - 1 {
+                if graphemes[idx].char == "d\u{361}" && graphemes[idx + 1].char == "ʒ" {
+                    let char_to_concat = graphemes[idx + 1].char.clone();
+                    graphemes[idx].char.push_str(&char_to_concat);
+                    graphemes.remove(idx + 1);
+                }
             }
         }
 

### 2023-04-14 paper.tex
Set date manually for latex
Since it thinks the date is January 1st 1980
Add initial parts of implementation for ot
Add recent changes, brush up deleting
Add introduction sec

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{minted}
 
 \usepackage{setspace}
-% \doublespacing
+\onehalfspacing
 
@@
 
-\title{Vowel Elision in Yoruba Collocations}
+\title{Computing Vowel Elision in Yoruba Collocations}
 \author{Hemant Gouni}
+\date{May 15th, 2023}
 
@@
 \maketitle
+
+\section{Introduction}
+
+The phenomenon of vowel hiatus in Yoruba, a language in the Niger-Congo family
+spoken primarily in West Africa, is well-documented. A large body of work
+exists that explores the various aspects of it, largely from an
+optimality-theoretic perspective \cite{ola2002yoruba, seidl2000yoruba}. Two
+particularly well-studied instances of this problem are \textit{assimilation}
+and \textit{deletion}. These are the two primary strategies employed by the
+language for resolving a sequence of two or more vowels that occur at the
+boundary of two adjoined morphemes. This text will specifically explore the
+second strategy of deletion, and in doing so, will attempt to provide an
+analysis that is simple enough to be feasible to verify computationally. That
+is, once we have developed our analysis, the core contribution of this work
+will be to encode it computationally so it can be quickly and easily verified
+whether it works for a range of corpus data. We employ an optimality-theoretic
+perspective throughout, and provide a thorough discussion in Section
+\ref{sec:computing-optimal-surface-forms} of the prior work on computational
+interpretations of optimality theory. Afterwards, we describe our own approach
+to automating our analysis and discuss its efficacy.
 
@@
     \textipa{/ow\'o-k\'i-ow\'o/} & \textipa{[ow\'ok\'ow\'o]} & \textipa{/\textdyoghlig{}\'o-\`Ew\`u/} & \textipa{[\textdyoghlig{}\'Ew\`u]}\\
-    \textipa{/OmO-k\'i-OmO/} & \textipa{[OmOk\'OmO]} & \textipa{/ra-\`Og\`Ed\`E/} & \textipa{[r\`Og\`Ed\`E]}\\
+    \textipa{/OmO-k\'i-OmO/} & \textipa{[OmOk\'OmO]} & \textipa{/ra-\`Og\`Ed\`E/} & \textipa{[r\`Og\`Ed\`E]}\footnotemark\\
     \textipa{/se-ol\'u/} & \textipa{[sol\'u]} & \textipa{/n\'i-oko/} & \textipa{[l\'oko]}\\
@@
 \end{figure}
+
+\footnotetext{As you will see, this example contradicts our analysis. However,
+it appears as a valid form only in a single source, so we note this and
+continue as though it is an exception.}
 
@@
 delete a morpheme-initial vowel over a more sonorous vowel (\textipa{[\'e]}
-rather than \textipa{[\'i]}). This ranking completely account for the pairs
+rather than \textipa{[\'i]}). This ranking completely accounts for the pairs
 shown in \pref{fig:deletion-except-1}. We would additionally like to rank
@@
 
+\label{sec:computing-optimal-surface-forms}
 \section{Computing Optimal Surface Forms}
@@
 surface form \textipa{[sol\'u]} and its underlying representation
-\textipa{/se-ol\'u/}. We will demonstrate our technique on this mapping.
+\textipa{/se-ol\'u/} \textbf{continue with this running example}. We will
+demonstrate our technique on this mapping.
 

### 2023-04-14 paper.tex
Discuss SyllabifiedCandidate, deletion implementation. Add impl for ONSET

@@
 \usepackage{minted}
+\usepackage{hyperref}
 
@@
 
-TODO: describe the algorithm succinctly using an algorithm figure?
+\subsection{Implementation}
 
-\subsection{Implementation}
+\begin{figure}[h]
+\caption{Our representation of words in Yoruba, along with our strategy for deletion.}
+\label{fig:impl-deletion}
+\begin{minted}{rust}
+#[derive(Debug, Clone)]
+struct SyllabifiedCandidate {
+    form: Vec<Segment>,
+    rng: StdRng,
+}
+
+impl SyllabifiedCandidate {
+    fn delete(mut self) -> Self {
+        if !self.form.is_empty() {
+            self.form.remove(self.rng.gen_range(0..self.form.len()));
+        }
+        self
+    }
+}
+\end{minted}
+\end{figure}
+
+We now describe key details of our implementation. The full source code may be
+found at \href{https://github.com/hemantgouni/linguistics-capstone}{github.com/hemantgouni/linguistics-capstone}.
+Starting with the basics, Figure \ref{fig:impl-deletion} shows both our
+representation of Yoruba words, and our simple handling of deletion.
+Specifically, a word is represented as a list of segments (this is the meaning of
+\texttt{Vec<Segment>}). At this point, you might ask: why not represent words simply
+as a list of characters? What's the difference between a character and a segment?
+This reflects a complication in computers' handling of text beyond simple English
+characters, covered by the ASCII character encoding. Unicode, an improvement of
+ASCII that handles many more languages beyond those which use the Latin alphabet,
+would consider the accent mark and letter \textipa{O} in \textipa{\'O}, to be
+separate characters, and not part of the same. We must sidestep this issue by
+parsing words into \textit{graphemes}, which accounts for the intuitive notion
+of what a character is.
+
+With that technical hiccup out of the way, our implementation of deletion is much
+simpler. It is interesting to note that within the context of our analysis,
+deletion turns out to be the most important of the operations to implement, because
+winning candidates are always generated via deletion (due to high-ranking \dep{}
+and \ident{} constraints). We implement deletion by picking a random element of
+the list of segments and removing it. Because we cluster characters by Unicode
+graphemes, the probability of any single character being deleted is evenly
+distributed between all in a given word.
+
+\begin{figure}[h]
+\caption{Our implementation of \onset}
+\label{fig:impl-onset}
+\begin{minted}{rust}
+impl Constraint for Onset {
+    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
+        let syllabi = surface
+            .form
+            .iter()
+            .filter(|seg| seg.syllable_index == SyllableIndex::Nucleus)
+            .count();
+
+        let onsets = surface
+            .form
+            .iter()
+            .filter(|seg| seg.syllable_index == SyllableIndex::Onset)
+            .count();
+
+        onsets - syllabi
+    }
+}
+\end{minted}
+\end{figure}
+
+We now
 

### 2023-04-14 paper.tex
Discuss Dep

@@
 
-\begin{figure}[h]
+\begin{figure}[h!]
 \caption{Our implementation of \onset}
 \label{fig:impl-onset}
-\begin{minted}{rust}
+\begin{minted}[linenos]{rust}
 impl Constraint for Onset {
@@
 
-We now
+The other significant portion of our implementation, beyond generating
+candidates, is evaluating them. This, of course, is done with constraints.
+Figure \ref{fig:impl-onset} shows our implementation of the \onset{}
+markedness constraint. On lines 2-7, we count the number of syllable
+nuclei in the input (a \texttt{SyllabifiedCandidate} from Figure
+\ref{fig:impl-deletion}). On lines 9-13, we count the number of onsets.
+Onsets and nuclei are calculated, along with other linguistic data, upon
+the creation of a \texttt{SyllabifiedCandidate}; we merely analyze them
+here. The number of violations of \onset{} is given by the number of
+onsets minus the number of nuclei. This will never be a negative number,
+because the algorithm that marks onsets and nuclei will only create one
+onset per nucleus, at most (because this is the maximum allowed by
+Yoruba).
 
-TODO: Figure out how best to implement the algorithm described above.
+\begin{figure}[h]
+\caption{The implementation of \dep}
+\label{fig:impl-dep}
+\begin{minted}[linenos]{rust}
+struct Dep(SyllabifiedCandidate);
 
-TODO: Show code examples of the critical parts, one for each of the
-subsections above!
+impl Constraint for Dep {
+    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
+        let self_str: String = self.0.into();
+        let surface_str: String = surface.into();
+
+        let diff = TextDiff::from_graphemes::<String>(&self_str, &surface_str);
+
+        diff.ops()
+            .iter()
+            .filter(|op| matches!(op, DiffOp::Insert { .. }))
+            .count()
+    }
+}
+\end{minted}
+\end{figure}
+
+Of course, no discussion of constraints would be complete without
+\textit{faithfulness} constraints. These present an interesting challenge,
+but we first provide an overview of the implementation, which is provided
+in Figure \ref{fig:impl-dep}. On lines 5 and 6, we respectively turn the
+original underlying form and the candidate surface form into their string
+respresentations. We then use a diff algorithm on these strings on line 8,
+which returns a sequence of operations which could be performed on the
+original underlying form to morph it into the candidate surface form given
+as input. On lines 10-13, we count the number of insertion operations
+returned by the diff algorithm.
+
+A difficult and subtle complication arises here: how does the diff
+algorithm \textit{know} what counts as an insertion, deletion, or change?
+For instance, changing \textipa{/ow\'o-k\'i-ow\'o/} into
+\textipa{[ow\'ok\'ow\'o]} is a relatively straightforward case of
+deletion. But what about \textipa{[ow\'ak\'iow\'o]}? In this case, the
+diff algorithm could decide that \textipa{[\'o]} was deleted and
+\textipa{[\'a]} inserted. Or it could instead say that \textipa{[\'o]}
+changed into \textipa{[\'a]}. Which of these occurs is linguistically
+relevant. This has not yet caused problems on real corpus data, but
+when it does, a potential solution is fine-tuning the diff algorithm
+to choose the more linguistically sound operations as outputs, making
+it aware of autosegmental phonology when it, for example, is choosing
+between a replacement operation or a insertion + deletion operations.
+
+To this point, our descriptions have been fairly abstract, so it is best
+to illustrate with an example. Given \textipa{/ow\'o-k\'i-ow\'o/} as
+input, how would we generate a random candidate output form from it?
+And how would we evaluate it against \onset{} and \dep?
 

### 2023-04-15 yoruba-ot/src/main.rs
Fix syllabification, Onset. Document coda issue

@@ -11,6 +11,9 @@ use utils::PushRet;
 // string -> syllabified candidate -> random deletions (all winners generated via deletions) ->
 // eval against constraints
 
+// TODO: make coda filling in more accurate to Yoruba, and make a constraint against unindexed
+// segments
+
 const VOWELS: [&str; 7] = ["o", "ɛ", "ɔ", "i", "u", "a", "e"];
 
 #[derive(Debug, Clone)]
@@ -45,7 +48,11 @@ impl SyllabifiedCandidate {
         if !self.form.is_empty() {
             self.form.remove(self.rng.gen_range(0..self.form.len()));
         }
-        self
+
+        SyllabifiedCandidate {
+            form: syllabify(self.form),
+            rng: self.rng,
+        }
     }
 }
 
@@ -208,7 +215,7 @@ impl Constraint for Onset {
             .filter(|seg| seg.syllable_index == SyllableIndex::Onset)
             .count();
 
-        onsets - syllabi
+        syllabi - onsets
     }
 }
 
@@ -234,8 +241,8 @@ impl Constraint for SonSeqPr {
 }
 
 fn main() {
-    let syllabified_candidate: SyllabifiedCandidate = dbg!("d͡ʒuigi".into());
-    dbg!(SonSeqPr.evaluate(syllabified_candidate));
+    let syllabified_candidate: SyllabifiedCandidate = dbg!("owokiowo".into());
+    dbg!(Onset.evaluate(syllabified_candidate));
 }
 
 #[cfg(test)]

### 2023-04-19 yoruba-ot/src/main.rs
Fix syllabification issue, not clearing indices

@@ -11,8 +11,9 @@ use utils::PushRet;
 // string -> syllabified candidate -> random deletions (all winners generated via deletions) ->
 // eval against constraints
 
-// TODO: make coda filling in more accurate to Yoruba, and make a constraint against unindexed
-// segments
+// TODO: make coda filling in more accurate to Yoruba (don't just replace all None with Coda, only
+// insert each coda at an available spot after each nucleus), and make a constraint against
+// unindexed segments
 
 const VOWELS: [&str; 7] = ["o", "ɛ", "ɔ", "i", "u", "a", "e"];
 
@@ -50,7 +51,22 @@ impl SyllabifiedCandidate {
         }
 
         SyllabifiedCandidate {
-            form: syllabify(self.form),
+            form: syllabify(self.clone().clear_indices().form),
+            rng: self.rng,
+        }
+    }
+
+    fn clear_indices(self) -> Self {
+        SyllabifiedCandidate {
+            form: self
+                .form
+                .iter()
+                .map(|seg| Segment {
+                    char: seg.char.clone(),
+                    syllable_index: SyllableIndex::None,
+                    seg_type: seg.seg_type.clone(),
+                })
+                .collect(),
             rng: self.rng,
         }
     }
@@ -241,8 +257,8 @@ impl Constraint for SonSeqPr {
 }
 
 fn main() {
-    let syllabified_candidate: SyllabifiedCandidate = dbg!("owokiowo".into());
-    dbg!(Onset.evaluate(syllabified_candidate));
+    let syllabified_candidate: SyllabifiedCandidate = dbg!("owoiowo".into());
+    dbg!(Onset.evaluate(dbg!(syllabified_candidate)));
 }
 
 #[cfg(test)]

### 2023-04-20 paper.tex

@@
 The phenomenon of vowel hiatus in Yoruba, a language in the Niger-Congo family
-spoken primarily in West Africa, is well-documented. A large body of work
+spoken primarily in West Africa, is well-documented. A mature body of work
 exists that explores the various aspects of it, largely from an
-optimality-theoretic perspective \cite{ola2002yoruba, seidl2000yoruba}. Two
-particularly well-studied instances of this problem are \textit{assimilation}
-and \textit{deletion}. These are the two primary strategies employed by the
-language for resolving a sequence of two or more vowels that occur at the
-boundary of two adjoined morphemes. This text will specifically explore the
-second strategy of deletion, and in doing so, will attempt to provide an
-analysis that is simple enough to be feasible to verify computationally. That
+optimality-theoretic perspective \cite{ola2002yoruba, seidl2000yoruba,
+pulleyblank1988vowel}. Two particularly well-studied instances of this problem
+are the \textit{assimilation} case and the \textit{deletion} case. These are the
+two primary strategies employed by the language for resolving sequences of two
+or more vowels that occur at the boundary of two adjoined morphemes
+(sequences of two vowels do not generally appear in Yoruba words due to
+syllable structure constraints). This text will specifically explore the
+second strategy of deletion, and in doing so, will specially attempt to provide
+an analysis that is simple enough to be feasible to verify computationally. That
 is, once we have developed our analysis, the core contribution of this work
-will be to encode it computationally so it can be quickly and easily verified
-whether it works for a range of corpus data. We employ an optimality-theoretic
-perspective throughout, and provide a thorough discussion in Section
-\ref{sec:computing-optimal-surface-forms} of the prior work on computational
-interpretations of optimality theory. Afterwards, we describe our own approach
-to automating our analysis and discuss its efficacy.
+will be to encode it as a computer program so it can be quickly and easily
+verified whether it works for a range of corpus data. We employ an
+optimality-theoretic perspective throughout, and provide a thorough discussion
+in Section \ref{sec:computing-optimal-surface-forms} of the prior work on
+computational interpretations of optimality theory. Afterwards, we describe
+our own approach to automating our analysis and discuss its efficacy.
 
@@
 the first vowel is deleted, unless the second vowel is less sonorous,
-in which case it is deleted. This deletion occurs in order to ensure that
-syllables always have an onset. This generalization can be analyzed from an
+in which case it is deleted (the Yoruba vowel inventory and sonority scale is
+discussed on page 4). This deletion occurs in order to ensure that syllables
+always have an onset. This generalization can be analyzed from an
 Optimality-theoretic viewpoint. To begin, the phenomena in
@@
 an onset. This is defined in \pref{def:onset-and-maxc}.
-
-% \textbf{make sure we're formulating the foot binarity effect correctly}
-
-% \textbf{cite Yoruba Vowel Elisision: Minimality Effects}
 
@@
 [i] over more sonorous vowels. We can codify this intuition into a faithfulness
-constraint, \ssp. This constraint is defined in \pref{def:ssp}.
+constraint, \ssp{} (which stands for `Sonority Sequencing Principle').
+This constraint is defined in \pref{def:ssp}.
 
@@
 
-We notice that candidate three is a winner threatening loser! It deletes both
+We notice that candidate four is a winner threatening loser! It deletes both
 of the non-sonorous vowels at the boundary of the two morphemes, and instead
@@
 
+\section{Computing Optimal Surface Forms}
 \label{sec:computing-optimal-surface-forms}
-\section{Computing Optimal Surface Forms}
 
@@
 Our first order of business is to generate a reasonably large set of
-permutations of \textipa{[sol\'u]}, one of which is the correct output form. We
-are not yet attempting to identify which is the correct output form, only
+permutations of \textipa{/se-ol\'u/}, one of which is the correct output form.
+We are not yet attempting to identify which is the correct output form, only
 generate it so it may be picked later. In effect, we are attempting to
@@
 \label{fig:impl-deletion}
-\begin{minted}{rust}
-#[derive(Debug, Clone)]
+\begin{minted}[linenos]{rust}
 struct SyllabifiedCandidate {
@@
 Starting with the basics, Figure \ref{fig:impl-deletion} shows both our
-representation of Yoruba words, and our simple handling of deletion.
-Specifically, a word is represented as a list of segments (this is the meaning of
-\texttt{Vec<Segment>}). At this point, you might ask: why not represent words simply
+representation of Yoruba words and our simple handling of deletion.
+Specifically, a word is represented as a list of segments; this is denoted by
+\texttt{Vec<Segment>} on line 2.
+At this point, it is reasonable to ask: why not represent words simply
 as a list of characters? What's the difference between a character and a segment?
-This reflects a complication in computers' handling of text beyond simple English
-characters, covered by the ASCII character encoding. Unicode, an improvement of
-ASCII that handles many more languages beyond those which use the Latin alphabet,
-would consider the accent mark and letter \textipa{O} in \textipa{\'O}, to be
-separate characters, and not part of the same. We must sidestep this issue by
+This reflects a complication in computers' handling of text beyond the
+simple English characters covered by the ASCII character encoding. Unicode, an
+improvement of ASCII that handles many more languages beyond those which use the
+Latin alphabet, would consider the accent mark and letter \textipa{O} in \textipa{\'O},
+to be separate characters, and not part of the same. We must sidestep this issue by
 parsing words into \textit{graphemes}, which accounts for the intuitive notion
-of what a character is.
+of what a character is: \textipa{\'O} is a single grapheme.
 
@@
 
-        onsets - syllabi
+        syllabi - onsets
     }
@@
 here. The number of violations of \onset{} is given by the number of
-onsets minus the number of nuclei. This will never be a negative number,
+nuclei minus the number of onsets. This will never be a negative number,
 because the algorithm that marks onsets and nuclei will only create one
@@
 
+For the sake of simplicity, we'll perform a single deletion operation.
+To do this, we must parse \textipa{/ow\'o-k\'i-ow\'o/} from a string
+into a \texttt{SyllabifiedCandidate}. We split it into each of its
+graphemes, creating the list \{\textipa{o,w,\'o,k,\'i,o,w,\'o}\}. To
+perform a deletion, we simply choose a random element of this list,
+say, [k]. This yields \{\textipa{o,w,\'o,\'i,o,w,\'o}\}.
+
+To find the number of \onset{} violations, we note that our list is
+annotated, internally, as \{\texttt{Nucleus}, \texttt{Onset},
+\texttt{Nucleus}, \texttt{Nucleus}, \texttt{Nucleus}, \texttt{Onset},
+\texttt{Nucleus}\}. Counting the number of nuclei, we find that there
+are 5. There are only two onsets. 5 - 2 = 3, so there are 3
+violations of \onset.
+
+Finally, to count the number of \dep{} violations, we turn
+\{\textipa{o,w,\'o,\'i,o,w,\'o}\} back into the string
+[\textipa{ow\'o\'iow\'o}]. We then use a diff algorithm to determine
+the sequence of edit operations that could be used to arrive at this
+string from \textipa{/ow\'o-k\'i-ow\'o/}. In this case, it tells us
+that a single \texttt{Delete} is all that is necessary, as expected.
+This does not cause any violations of \dep{} because it is not an
+\texttt{Insert} operation, so we return 0 violations.
+
 \subsection{Analysis on corpus data}
@@
 
-TODO: Requires implementation to be reasonably far along first!
+TODO: Do the evaluations now that the implementation is reasonably
+complete!
 

### 2023-04-20 yoruba-ot/src/main.rs
Resolve TODO about fixing coda marking

@@ -11,10 +11,6 @@ use utils::PushRet;
 // string -> syllabified candidate -> random deletions (all winners generated via deletions) ->
 // eval against constraints
 
-// TODO: make coda filling in more accurate to Yoruba (don't just replace all None with Coda, only
-// insert each coda at an available spot after each nucleus), and make a constraint against
-// unindexed segments
-
 const VOWELS: [&str; 7] = ["o", "ɛ", "ɔ", "i", "u", "a", "e"];
 
 #[derive(Debug, Clone)]
@@ -168,11 +164,21 @@ fn mark_onsets(candidate: Vec<Segment>) -> Vec<Segment> {
 fn mark_codas(candidate: Vec<Segment>) -> Vec<Segment> {
     candidate
         .iter()
-        .map(|seg| match seg.syllable_index {
-            SyllableIndex::None => seg.morph_type(SyllableIndex::Coda),
-            _ => seg.to_owned(),
-        })
-        .collect()
+        .fold(
+            (SyllableIndex::None, Vec::new()),
+            |(prev_seg_type, mut segs), seg| match (prev_seg_type, seg.syllable_index.to_owned()) {
+                (SyllableIndex::Nucleus, SyllableIndex::None) => (
+                    SyllableIndex::Coda,
+                    segs.push_ret(seg.morph_type(SyllableIndex::Coda))
+                        .to_owned(),
+                ),
+                _ => (
+                    seg.syllable_index.to_owned(),
+                    segs.push_ret(seg.to_owned()).to_owned(),
+                ),
+            },
+        )
+        .1
 }
 
 fn syllabify(candidate: Vec<Segment>) -> Vec<Segment> {
@@ -256,8 +262,20 @@ impl Constraint for SonSeqPr {
     }
 }
 
+struct Syllabify;
+
+impl Constraint for Syllabify {
+    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
+        surface
+            .form
+            .iter()
+            .filter(|seg| seg.syllable_index == SyllableIndex::None)
+            .count()
+    }
+}
+
 fn main() {
-    let syllabified_candidate: SyllabifiedCandidate = dbg!("owoiowo".into());
+    let syllabified_candidate: SyllabifiedCandidate = dbg!("owoktwiowo".into());
     dbg!(Onset.evaluate(dbg!(syllabified_candidate)));
 }
 

### 2023-04-20 paper.tex
Updates from Overleaf

@@
 Latin alphabet, would consider the accent mark and letter \textipa{O} in \textipa{\'O},
-to be separate characters, and not part of the same. We must sidestep this issue by
+to be separate characters and not parts of the same. We must sidestep this issue by
 parsing words into \textit{graphemes}, which accounts for the intuitive notion

### 2023-04-20 yoruba-ot/src/constraint.rs
Split out constraints into their own module

@@ -1,0 +1,91 @@
+use crate::{SyllabifiedCandidate, SyllableIndex};
+use similar::{DiffOp, TextDiff};
+
+pub trait Constraint {
+    fn evaluate(self, surface: SyllabifiedCandidate) -> usize;
+}
+
+pub struct Ident(pub SyllabifiedCandidate);
+
+impl Constraint for Ident {
+    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
+        let self_str: String = self.0.into();
+        let surface_str: String = surface.into();
+
+        let diff = TextDiff::from_graphemes::<String>(&self_str, &surface_str);
+
+        diff.ops()
+            .iter()
+            .filter(|op| matches!(op, DiffOp::Replace { .. }))
+            .count()
+    }
+}
+
+pub struct Dep(pub SyllabifiedCandidate);
+
+impl Constraint for Dep {
+    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
+        let self_str: String = self.0.into();
+        let surface_str: String = surface.into();
+
+        let diff = TextDiff::from_graphemes::<String>(&self_str, &surface_str);
+
+        diff.ops()
+            .iter()
+            .filter(|op| matches!(op, DiffOp::Insert { .. }))
+            .count()
+    }
+}
+
+pub struct Onset;
+
+impl Constraint for Onset {
+    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
+        let syllabi = surface
+            .form
+            .iter()
+            .filter(|seg| seg.syllable_index == SyllableIndex::Nucleus)
+            .count();
+
+        let onsets = surface
+            .form
+            .iter()
+            .filter(|seg| seg.syllable_index == SyllableIndex::Onset)
+            .count();
+
+        syllabi - onsets
+    }
+}
+
+pub struct SonSeqPr;
+
+impl Constraint for SonSeqPr {
+    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
+        surface
+            .form
+            .iter()
+            // a hack to ignore accent marks
+            //
+            // .next().unwrap() should never panic here bc that's only possible if the initial
+            // candidate input string is empty, and if that's true, then the iterator will be empty
+            .map(|seg| match dbg!(seg.char.chars().next().unwrap()) {
+                'e' | 'ɛ' | 'o' | 'ɔ' => 1,
+                'u' => 2,
+                'i' => 3,
+                _ => 0,
[…]

### 2023-04-20 yoruba-ot/src/main.rs

@@ -1,9 +1,10 @@
-#![allow(dead_code)]
+#![allow(unused_imports)]
 
+mod constraint;
 mod utils;
 
+use crate::constraint::{Constraint, Dep, Ident, Onset, SonSeqPr, Syllabify};
 use rand::{rngs::StdRng, Rng, SeedableRng};
-use similar::{DiffOp, TextDiff};
 use unicode_segmentation::UnicodeSegmentation;
 
 use utils::PushRet;
@@ -14,7 +15,7 @@ use utils::PushRet;
 const VOWELS: [&str; 7] = ["o", "ɛ", "ɔ", "i", "u", "a", "e"];
 
 #[derive(Debug, Clone)]
-struct SyllabifiedCandidate {
+pub struct SyllabifiedCandidate {
     form: Vec<Segment>,
     rng: StdRng,
 }
@@ -185,98 +186,10 @@ fn syllabify(candidate: Vec<Segment>) -> Vec<Segment> {
     mark_codas(mark_onsets(mark_vowels(candidate)))
 }
 
-trait Constraint {
-    fn evaluate(self, surface: SyllabifiedCandidate) -> usize;
-}
-
-struct Ident(SyllabifiedCandidate);
-
-impl Constraint for Ident {
-    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
-        let self_str: String = self.0.into();
-        let surface_str: String = surface.into();
-
-        let diff = TextDiff::from_graphemes::<String>(&self_str, &surface_str);
-
-        diff.ops()
-            .iter()
-            .filter(|op| matches!(op, DiffOp::Replace { .. }))
-            .count()
-    }
-}
-
-struct Dep(SyllabifiedCandidate);
-
-impl Constraint for Dep {
-    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
-        let self_str: String = self.0.into();
-        let surface_str: String = surface.into();
-
-        let diff = TextDiff::from_graphemes::<String>(&self_str, &surface_str);
-
-        diff.ops()
-            .iter()
-            .filter(|op| matches!(op, DiffOp::Insert { .. }))
-            .count()
-    }
-}
-
-struct Onset;
-
-impl Constraint for Onset {
-    fn evaluate(self, surface: SyllabifiedCandidate) -> usize {
-        let syllabi = surface
-            .form
-            .iter()
-            .filter(|seg| seg.syllable_index == SyllableIndex::Nucleus)
-            .count();
-
[…]

### 2023-05-01 yoruba-ot/src/main.rs
Move Syllabify import, only used in tests

@@ -3,7 +3,6 @@ mod utils;
 
 use crate::constraint::{
     Constraint, Dep, Ident, Max, MaxFinalV, MaxInitialV, Onset, RankedConstraint, SonSeqPr,
-    Syllabify,
 };
 use itertools::Itertools;
 use unicode_segmentation::UnicodeSegmentation;
@@ -357,6 +356,8 @@ fn main() {
 #[cfg(test)]
 mod test {
     use super::*;
+    use crate::constraint::Syllabify;
+
     #[test]
     fn test_ident_1() {
         let cand1: SyllabifiedCandidate = "owókíowó".into();

### 2023-05-07 paper.tex
Finish analysis, write conclusion (rough)

@@
 \section{Introduction}
+
+% Maybe add some background as Yoruba
 
@@
 
-\begin{figure}[h]
-\caption{Yoruba Surface Forms}
-\label{fig:data}
-\begin{tabular}{c c c c}
-    \textbf{Word} & \textbf{Gloss} & \textbf{Word} & \textbf{Gloss}\\
-    \textipa{ow\'o} & money & \textipa{k\'i} & any\\
-    \textipa{OmO} & child & se & cook\\
-    \textipa{ol\'u} & mushrooms & \textipa{\textdyoghlig{}E} & eat\\
-    \textipa{ed\'e} & shrimp & \textipa{\textdyoghlig{}\'o} & burn\\
-    \textipa{\`Ew\`u} & clothing & \textipa{Ra} & buy\\
-    \textipa{\`Og\`Ed\`E} & banana & \textipa{n\'i} & at\\
-    \textipa{s\'i} & to & \textipa{O\textdyoghlig{}\`a} & market\\
-    \textipa{ow\'ok\'ow\'o} & any money at all | bad money & \textipa{\textdyoghlig{}ed\'e} & eat shrimp\\
-    \textipa{OmOk\'OmO} & any child at all | bad child & \textipa{sol\'u} & cook mushrooms\\
-    \textipa{\textdyoghlig{}\'Ew\`u} & burn clothing & \textipa{R\`Og\`Ed\`E} & buy bananas\\
-    \textipa{l\'oko} & at the farm & \textipa{s\'O\textdyoghlig{}\`a} & to the market\\
-    \textipa{gb\'e} & lift & \textipa{in\'O} & lamp\\
-    \textipa{gba} & take & \textipa{iS\'E} & job\\
-    \textipa{wo} & look & \textipa{il\`E} & ground\\
-    \textipa{\textdyoghlig{}\'i} & steal & \textipa{aSO} & clothes\\
-    \textipa{gb\'en\'O} & lift the lamp & \textipa{gbaS\'E} & take a job\\
-    \textipa{wol\`E} & look at the ground & \textipa{\textdyoghlig{}\'aSO} & steal clothes\\
-    \textipa{\textdyoghlig{}u} & throw & \textipa{igi} & stick\\
-    \textipa{lu} & hit & \textipa{il\`E} & ground\\
-    \textipa{bu} & take & \textipa{omi} & water\\
-    \textipa{Ru} & carry & \textipa{epo} & oil\\
-    \textipa{b\'u} & abuse & \textipa{Oba} & king\\
-    \textipa{\textdyoghlig{}ugi} & throw a stick & \textipa{lul\`E} & hit the ground\\
-    \textipa{bomi} & take water & \textipa{Repo} & carry oil\\
-    \textipa{b\'Oba} & abuse the king & \textipa{b\'i} & have\\
-    \textipa{b\'imO} & have a baby\\
-\end{tabular}
-\end{figure}
-
-\pagebreak
-
 \section{Computing Optimal Surface Forms}
@@
 \subsection{Approach}
+\label{subsec:approach}
 
 We base our procedure, crucially, on bounded depth-first search. We first
-describe the basic algorithm, before proceeding onwards to an implementation.
-Viewed as a single function which takes an input and returns an output, our
-analysis takes as input an underlying form and a set of ranked constraints,
-and searches for the correct output form. The search for the correct output
-is of course the crucial portion here, and is the part on which most of the
+describe a basic and idealized algorithm, before proceeding onwards to an
+implementation section where we will discuss certain shortcuts which can be
+taken due to convenient properties of vowel elision in Yoruba. Viewed as a
+single function which takes an input and returns an output, our analysis 
+takes as input an underlying form and a set of ranked constraints, and
+searches for the correct output form. The search for the correct output is
+of course the crucial portion here, and is the part on which most of the
 discussion will be spent.
@@
 found at \href{https://github.com/hemantgouni/linguistics-capstone}{github.com/hemantgouni/linguistics-capstone}.
-Starting with the basics, Figure \ref{fig:impl-deletion} shows both our
+Starting with the basics, Figure \pref{fig:impl-deletion} shows both our
 representation of Yoruba words and our simple handling of deletion.
@@
 winning candidates are always generated via deletion (due to high-ranking \dep{}
-and \ident{} constraints). We implement deletion by picking a random element of
-the list of segments and removing it. Because we cluster characters by Unicode
-graphemes, the probability of any single character being deleted is evenly
-distributed between all in a given word.
+and \ident{} constraints). In fact, in the context of our specific analysis, it is
+easy to generate \textit{all possible} surface forms instead of limiting ourselves to
+the previously mentioned 65536 output forms, simply by choosing to delete or not
+delete each segment. This yields $2^n$ possible output forms for a given input
+form, where $n$ is the number of segments in the input form. We implement deletion
+by picking a random element of the list of segments and removing it. Because we
+cluster characters by Unicode graphemes, the probability of any single character
+being deleted is evenly distributed between all in a given word.
 
@@
 candidates, is evaluating them. This, of course, is done with constraints.
-Figure \ref{fig:impl-onset} shows our implementation of the \onset{}
+Figure \pref{fig:impl-onset} shows our implementation of the \onset{}
 markedness constraint. On lines 2-7, we count the number of syllable
 nuclei in the input (a \texttt{SyllabifiedCandidate} from Figure
-\ref{fig:impl-deletion}). On lines 9-13, we count the number of onsets.
+\pref{fig:impl-deletion}). On lines 9-13, we count the number of onsets.
 Onsets and nuclei are calculated, along with other linguistic data, upon
@@
 but we first provide an overview of the implementation, which is provided
-in Figure \ref{fig:impl-dep}. On lines 5 and 6, we respectively turn the
+in Figure \pref{fig:impl-dep}. On lines 5 and 6, we respectively turn the
 original underlying form and the candidate surface form into their string
@@
 In this section we will evaluate all of our collected Yoruba underlying
-representations against the expected surface forms.
+representations against the expected surface forms. The constraints
+used in our analysis, along with their rankings, are listed here:
 
-TODO: Do the evaluations now that the implementation is reasonably
-complete!
+\begin{enumerate}
+    \item
+        \begin{enumerate}
+            \item \ident{}
+            \item \dep{}
+        \end{enumerate}
+    \item
+        \begin{enumerate}
+            \item \onset{}
+            \item \ssp{}
+            \item \maxc{}
+        \end{enumerate}
+    \item
+        \begin{enumerate}
+            \item \maxplusv{}
+        \end{enumerate}
+    \item
+        \begin{enumerate}
+            \item \maxvplus{}
+        \end{enumerate}
+\end{enumerate}
+
+The number under which each constraint is listed denotes its rank.
+For example, \ident{} and \dep{} have the highest ranking, followed
+by \onset{}, \ssp{}, and \maxc{}. This largely matches
+Fig. \pref{tableau:summary-two}, but with the inclusion of an extra
+\maxc{} constraint at rank two. This is necessary because,
+otherwise, our program (correctly) finds deleting the input form
+to be the optimal solution to the set of constraints given. That is,
+neither of the rank 1 constraints (\ident{} and \dep{}), nor the
+constraints immediately below them (\onset{} and \ssp{}), generate
+violations for deleting words. On the other hand, a form with no
+segments will \textit{always} generate less violations of \onset{}
+and \ssp{} than one which does-- the mere presence of a vowel in a
+form is sufficient to generate \ssp{} violations. Adding a \maxc{}
+constraint ranked equally to \onset{} and \ssp{} addresses this.
+Indeed, this is not an issue with our implementation, but a more
+fundamental issue with our analysis which was found by it. In that
+sense, we can see that the benefits of computationally encoding the
+analysis are already emerging.
+
+An additional point to note is that, when evaluating candidates,
+instead of considering equally-ranked constraints individually, we
+treat violations between them cumulatively. As a result, since
+\ssp{} generates multiple violations at a time (for instance [i]
+will generate 3 violations, per Figure
+\pref{table:vowel-sonority-v2}), in order for \maxc{} to
+adequately regulate against deletion, it must generate multiple
+(specifically, 3) violations for each deletion. In addition,
+\onset{}, being equally ranked to \ssp{} and \maxc{}, must also
+do this. This quirk of our implementation is certainly
+unconventional among the OT literature, and further study might
+better support some of these choices (or provide more elegant
+alternative solutions).
+
+Beyond these issue, our analysis is sound. Given an underlying
+Yoruba form, the program implementing it will produce the
+appropriately vowel-elided surface form. In other words, given
+the information in the `Underlying' columns of tables
+\pref{fig:deletion-simple}, \pref{fig:deletion-except-1},
+\pref{fig:deletion-except-1}, our program outputs the
+form expected in the 'Surface' columns (excepting
+\textipa{/n\'i-oko/} \rightarrow{} \textipa{[l\'oko]},
+since the n \rightarrow{} l transition occurs for reasons
+unrelated to our analysis; we also ignore tone markers
+here, because it is a related but orthogonal phenomenon
+unhandled by our original analysis).
+
+In fact, it will even produce the intuitively expected surface
+form when given underlying forms which do not exist. For
+instance, given \textipa{[so-el\'u]} as an input, it will produce
+\textipa{/sel\'u/}, due to \maxplusv{} regulating more
+strictly against initial vowel deletion than deletion of
+final vowels.
 
 \section{Concluding}
+
+In this paper, we presented a classic Optimality-Theoretic
+analysis of conjoined vowel elision in Yoruba. We then
+implemented it as a computer program, finding a subtle
+issue in the analysis. With these issues addressed in the
+implementation, our program outputs the correctly
+vowel-elided surface form, given an underlying
+representation. Future work could go in several interesting
+directions. Within the domain of Yoruba, the original
+analysis and its implementation could be extended to handle
+the assimilatory strategy for dealing with sequences of
+vowels at morpheme boundaries (as discussed in
+\cite{ola2002yoruba}), as well as tone. This would likely
+involve extending surface candidate generation to the full
+idealized algorithm described in Section
+\ref{subsec:approach}, contending with epenthesis,
+metathesis, and segment changes in addition to deletion.
+The implementation, as it exists, faithfully models the
+parts of the algorithm in \ref{subsec:approach} that are
+necessary in the context of Yoruba, so this would largely
+be a case of extending the existing machinery.
+
+\pagebreak
+
+\section{Appendix}
+
+\begin{figure}[h]
+\caption{Yoruba Surface Forms}
+\label{fig:data}
+\begin{tabular}{c c c c}
+    \textbf{Word} & \textbf{Gloss} & \textbf{Word} & \textbf{Gloss}\\
+    \textipa{ow\'o} & money & \textipa{k\'i} & any\\
+    \textipa{OmO} & child & se & cook\\
+    \textipa{ol\'u} & mushrooms & \textipa{\textdyoghlig{}E} & eat\\
+    \textipa{ed\'e} & shrimp & \textipa{\textdyoghlig{}\'o} & burn\\
+    \textipa{\`Ew\`u} & clothing & \textipa{Ra} & buy\\
+    \textipa{\`Og\`Ed\`E} & banana & \textipa{n\'i} & at\\
+    \textipa{s\'i} & to & \textipa{O\textdyoghlig{}\`a} & market\\
+    \textipa{ow\'ok\'ow\'o} & any money at all | bad money & \textipa{\textdyoghlig{}ed\'e} & eat shrimp\\
+    \textipa{OmOk\'OmO} & any child at all | bad child & \textipa{sol\'u} & cook mushrooms\\
+    \textipa{\textdyoghlig{}\'Ew\`u} & burn clothing & \textipa{R\`Og\`Ed\`E} & buy bananas\\
+    \textipa{l\'oko} & at the farm & \textipa{s\'O\textdyoghlig{}\`a} & to the market\\
+    \textipa{gb\'e} & lift & \textipa{in\'O} & lamp\\
+    \textipa{gba} & take & \textipa{iS\'E} & job\\
+    \textipa{wo} & look & \textipa{il\`E} & ground\\
+    \textipa{\textdyoghlig{}\'i} & steal & \textipa{aSO} & clothes\\
+    \textipa{gb\'en\'O} & lift the lamp & \textipa{gbaS\'E} & take a job\\
+    \textipa{wol\`E} & look at the ground & \textipa{\textdyoghlig{}\'aSO} & steal clothes\\
+    \textipa{\textdyoghlig{}u} & throw & \textipa{igi} & stick\\
+    \textipa{lu} & hit & \textipa{il\`E} & ground\\
+    \textipa{bu} & take & \textipa{omi} & water\\
+    \textipa{Ru} & carry & \textipa{epo} & oil\\
+    \textipa{b\'u} & abuse & \textipa{Oba} & king\\
+    \textipa{\textdyoghlig{}ugi} & throw a stick & \textipa{lul\`E} & hit the ground\\
+    \textipa{bomi} & take water & \textipa{Repo} & carry oil\\
+    \textipa{b\'Oba} & abuse the king & \textipa{b\'i} & have\\
+    \textipa{b\'imO} & have a baby\\
+\end{tabular}
+\end{figure}
+
+\pagebreak
 

### 2023-05-07 paper.tex
Address peer review comments

@@
 morpheme-final positions. We define both of these constraints in
-\pref{def:maxvplus-and-maxplusv}.
+\pref{def:maxvplus-and-maxplusv}. Note that we could define these without
+specifying \textsc{-V}; that is, they could simply regulate against deletion
+at morpheme boundaries, but we have chosen to specialize these to our analysis
+to avoid unintended interactions with deletion in Yoruba unrelated to our work.
 
@@
 \pref{fig:deletion-except-2} by ranking /i/ as lower in the sonority scale of
-Yoruba. This elegantly deals with the issue of /i/ deletion at the beginning of
+Yoruba. This deals with the issue of /i/ deletion at the beginning of
 the second vowel being preferred over /u/ deletion at the end of the first
@@
 theorizing that Yoruba possesses a specialized variant of it, as many languages
-do.
+do. This sonority scale is motivated only linguistically, within the context
+of our analysis, and not psycho-acoustically-- though that would be an
+interesting direction for future work.
 

### 2023-05-07 paper.tex
Fix line breaking
Since Overleaf's Vi mode doesn't implement it

@@
 We now describe key details of our implementation. The full source code may be
-found at \href{https://github.com/hemantgouni/linguistics-capstone}{github.com/hemantgouni/linguistics-capstone}.
+found at
+\href{https://github.com/hemantgouni/linguistics-capstone}{github.com/hemantgouni/linguistics-capstone}.
 Starting with the basics, Figure \pref{fig:impl-deletion} shows both our
@@
 Specifically, a word is represented as a list of segments; this is denoted by
-\texttt{Vec<Segment>} on line 2.
-At this point, it is reasonable to ask: why not represent words simply
-as a list of characters? What's the difference between a character and a segment?
-This reflects a complication in computers' handling of text beyond the
-simple English characters covered by the ASCII character encoding. Unicode, an
-improvement of ASCII that handles many more languages beyond those which use the
-Latin alphabet, would consider the accent mark and letter \textipa{O} in \textipa{\'O},
-to be separate characters and not parts of the same. We must sidestep this issue by
-parsing words into \textit{graphemes}, which accounts for the intuitive notion
-of what a character is: \textipa{\'O} is a single grapheme.
+\texttt{Vec<Segment>} on line 2. At this point, it is reasonable to ask: why
+not represent words simply as a list of characters? What's the difference
+between a character and a segment? This reflects a complication in computers'
+handling of text beyond the simple English characters covered by the ASCII
+character encoding. Unicode, an improvement of ASCII that handles many more
+languages beyond those which use the Latin alphabet, would consider the accent
+mark and letter \textipa{O} in \textipa{\'O}, to be separate characters and not
+parts of the same. We must sidestep this issue by parsing words into
+\textit{graphemes}, which accounts for the intuitive notion of what a character
+is: \textipa{\'O} is a single grapheme.
 
-With that technical hiccup out of the way, our implementation of deletion is much
-simpler. It is interesting to note that within the context of our analysis,
-deletion turns out to be the most important of the operations to implement, because
-winning candidates are always generated via deletion (due to high-ranking \dep{}
-and \ident{} constraints). In fact, in the context of our specific analysis, it is
-easy to generate \textit{all possible} surface forms instead of limiting ourselves to
-the previously mentioned 65536 output forms, simply by choosing to delete or not
-delete each segment. This yields $2^n$ possible output forms for a given input
-form, where $n$ is the number of segments in the input form. We implement deletion
-by picking a random element of the list of segments and removing it. Because we
-cluster characters by Unicode graphemes, the probability of any single character
-being deleted is evenly distributed between all in a given word.
+With that technical hiccup out of the way, our implementation of deletion is
+much simpler. It is interesting to note that within the context of our
+analysis, deletion turns out to be the most important of the operations to
+implement, because winning candidates are always generated via deletion (due to
+high-ranking \dep{} and \ident{} constraints). In fact, in the context of our
+specific analysis, it is easy to generate \textit{all possible} surface forms
+instead of limiting ourselves to the previously mentioned 65536 output forms,
+simply by choosing to delete or not delete each segment. This yields $2^n$
+possible output forms for a given input form, where $n$ is the number of
+segments in the input form. We implement deletion by picking a random element
+of the list of segments and removing it. Because we cluster characters by
+Unicode graphemes, the probability of any single character being deleted is
+evenly distributed between all in a given word.
 
@@
 Figure \pref{fig:impl-onset} shows our implementation of the \onset{}
-markedness constraint. On lines 2-7, we count the number of syllable
-nuclei in the input (a \texttt{SyllabifiedCandidate} from Figure
-\pref{fig:impl-deletion}). On lines 9-13, we count the number of onsets.
-Onsets and nuclei are calculated, along with other linguistic data, upon
-the creation of a \texttt{SyllabifiedCandidate}; we merely analyze them
-here. The number of violations of \onset{} is given by the number of
-nuclei minus the number of onsets. This will never be a negative number,
-because the algorithm that marks onsets and nuclei will only create one
-onset per nucleus, at most (because this is the maximum allowed by
-Yoruba).
+markedness constraint. On lines 2-7, we count the number of syllable nuclei in
+the input (a \texttt{SyllabifiedCandidate} from Figure
+\pref{fig:impl-deletion}). On lines 9-13, we count the number of onsets. Onsets
+and nuclei are calculated, along with other linguistic data, upon the creation
+of a \texttt{SyllabifiedCandidate}; we merely analyze them here. The number of
+violations of \onset{} is given by the number of nuclei minus the number of
+onsets. This will never be a negative number, because the algorithm that marks
+onsets and nuclei will only create one onset per nucleus, at most (because this
+is the maximum allowed by Yoruba).
 
@@
 Of course, no discussion of constraints would be complete without
-\textit{faithfulness} constraints. These present an interesting challenge,
-but we first provide an overview of the implementation, which is provided
-in Figure \pref{fig:impl-dep}. On lines 5 and 6, we respectively turn the
-original underlying form and the candidate surface form into their string
+\textit{faithfulness} constraints. These present an interesting challenge, but
+we first provide an overview of the implementation, which is provided in Figure
+\pref{fig:impl-dep}. On lines 5 and 6, we respectively turn the original
+underlying form and the candidate surface form into their string
 respresentations. We then use a diff algorithm on these strings on line 8,
-which returns a sequence of operations which could be performed on the
-original underlying form to morph it into the candidate surface form given
-as input. On lines 10-13, we count the number of insertion operations
-returned by the diff algorithm.
+which returns a sequence of operations which could be performed on the original
+underlying form to morph it into the candidate surface form given as input. On
+lines 10-13, we count the number of insertion operations returned by the diff
+algorithm.
 
-A difficult and subtle complication arises here: how does the diff
-algorithm \textit{know} what counts as an insertion, deletion, or change?
-For instance, changing \textipa{/ow\'o-k\'i-ow\'o/} into
-\textipa{[ow\'ok\'ow\'o]} is a relatively straightforward case of
-deletion. But what about \textipa{[ow\'ak\'iow\'o]}? In this case, the
-diff algorithm could decide that \textipa{[\'o]} was deleted and
-\textipa{[\'a]} inserted. Or it could instead say that \textipa{[\'o]}
-changed into \textipa{[\'a]}. Which of these occurs is linguistically
-relevant. This has not yet caused problems on real corpus data, but
-when it does, a potential solution is fine-tuning the diff algorithm
-to choose the more linguistically sound operations as outputs, making
-it aware of autosegmental phonology when it, for example, is choosing
-between a replacement operation or a insertion + deletion operations.
+A difficult and subtle complication arises here: how does the diff algorithm
+\textit{know} what counts as an insertion, deletion, or change? For instance,
+changing \textipa{/ow\'o-k\'i-ow\'o/} into \textipa{[ow\'ok\'ow\'o]} is a
+relatively straightforward case of deletion. But what about
+\textipa{[ow\'ak\'iow\'o]}? In this case, the diff algorithm could decide that
+\textipa{[\'o]} was deleted and \textipa{[\'a]} inserted. Or it could instead
+say that \textipa{[\'o]} changed into \textipa{[\'a]}. Which of these occurs is
+linguistically relevant. This has not yet caused problems on real corpus data,
+but when it does, a potential solution is fine-tuning the diff algorithm to
+choose the more linguistically sound operations as outputs, making it aware of
+autosegmental phonology when it, for example, is choosing between a replacement
+operation or a insertion + deletion operations.
 
-To this point, our descriptions have been fairly abstract, so it is best
-to illustrate with an example. Given \textipa{/ow\'o-k\'i-ow\'o/} as
-input, how would we generate a random candidate output form from it?
-And how would we evaluate it against \onset{} and \dep?
+To this point, our descriptions have been fairly abstract, so it is best to
+illustrate with an example. Given \textipa{/ow\'o-k\'i-ow\'o/} as input, how
+would we generate a random candidate output form from it? And how would we
+evaluate it against \onset{} and \dep?
 
-For the sake of simplicity, we'll perform a single deletion operation.
-To do this, we must parse \textipa{/ow\'o-k\'i-ow\'o/} from a string
-into a \texttt{SyllabifiedCandidate}. We split it into each of its
-graphemes, creating the list \{\textipa{o,w,\'o,k,\'i,o,w,\'o}\}. To
-perform a deletion, we simply choose a random element of this list,
-say, [k]. This yields \{\textipa{o,w,\'o,\'i,o,w,\'o}\}.
+For the sake of simplicity, we'll perform a single deletion operation. To do
+this, we must parse \textipa{/ow\'o-k\'i-ow\'o/} from a string into a
+\texttt{SyllabifiedCandidate}. We split it into each of its graphemes, creating
+the list \{\textipa{o,w,\'o,k,\'i,o,w,\'o}\}. To perform a deletion, we simply
+choose a random element of this list, say, [k]. This yields
+\{\textipa{o,w,\'o,\'i,o,w,\'o}\}.
 
-To find the number of \onset{} violations, we note that our list is
-annotated, internally, as \{\texttt{Nucleus}, \texttt{Onset},
-\texttt{Nucleus}, \texttt{Nucleus}, \texttt{Nucleus}, \texttt{Onset},
-\texttt{Nucleus}\}. Counting the number of nuclei, we find that there
-are 5. There are only two onsets. 5 - 2 = 3, so there are 3
-violations of \onset.
+To find the number of \onset{} violations, we note that our list is annotated,
+internally, as \{\texttt{Nucleus}, \texttt{Onset}, \texttt{Nucleus},
+\texttt{Nucleus}, \texttt{Nucleus}, \texttt{Onset}, \texttt{Nucleus}\}.
+Counting the number of nuclei, we find that there are 5. There are only two
+onsets. 5 - 2 = 3, so there are 3 violations of \onset.
 
@@
 \{\textipa{o,w,\'o,\'i,o,w,\'o}\} back into the string
-[\textipa{ow\'o\'iow\'o}]. We then use a diff algorithm to determine
-the sequence of edit operations that could be used to arrive at this
-string from \textipa{/ow\'o-k\'i-ow\'o/}. In this case, it tells us
-that a single \texttt{Delete} is all that is necessary, as expected.
-This does not cause any violations of \dep{} because it is not an
-\texttt{Insert} operation, so we return 0 violations.
+[\textipa{ow\'o\'iow\'o}]. We then use a diff algorithm to determine the
+sequence of edit operations that could be used to arrive at this string from
+\textipa{/ow\'o-k\'i-ow\'o/}. In this case, it tells us that a single
+\texttt{Delete} is all that is necessary, as expected. This does not cause any
+violations of \dep{} because it is not an \texttt{Insert} operation, so we
+return 0 violations.
 
@@
 In this section we will evaluate all of our collected Yoruba underlying
-representations against the expected surface forms. The constraints
-used in our analysis, along with their rankings, are listed here:
+representations against the expected surface forms. The constraints used in our
+analysis, along with their rankings, are listed here:
 
@@
 
-The number under which each constraint is listed denotes its rank.
-For example, \ident{} and \dep{} have the highest ranking, followed
-by \onset{}, \ssp{}, and \maxc{}. This largely matches
-Fig. \pref{tableau:summary-two}, but with the inclusion of an extra
-\maxc{} constraint at rank two. This is necessary because,
-otherwise, our program (correctly) finds deleting the input form
-to be the optimal solution to the set of constraints given. That is,
-neither of the rank 1 constraints (\ident{} and \dep{}), nor the
-constraints immediately below them (\onset{} and \ssp{}), generate
-violations for deleting words. On the other hand, a form with no
-segments will \textit{always} generate less violations of \onset{}
-and \ssp{} than one which does-- the mere presence of a vowel in a
-form is sufficient to generate \ssp{} violations. Adding a \maxc{}
-constraint ranked equally to \onset{} and \ssp{} addresses this.
-Indeed, this is not an issue with our implementation, but a more
-fundamental issue with our analysis which was found by it. In that
-sense, we can see that the benefits of computationally encoding the
-analysis are already emerging.
+The number under which each constraint is listed denotes its rank. For example,
+\ident{} and \dep{} have the highest ranking, followed by \onset{}, \ssp{}, and
+\maxc{}. This largely matches Fig. \pref{tableau:summary-two}, but with the
+inclusion of an extra \maxc{} constraint at rank two. This is necessary
+because, otherwise, our program (correctly) finds deleting the input form to be
+the optimal solution to the set of constraints given. That is, neither of the
+rank 1 constraints (\ident{} and \dep{}), nor the constraints immediately below
+them (\onset{} and \ssp{}), generate violations for deleting words. On the
+other hand, a form with no segments will \textit{always} generate less
+violations of \onset{} and \ssp{} than one which does-- the mere presence of a
+vowel in a form is sufficient to generate \ssp{} violations. Adding a \maxc{}
+constraint ranked equally to \onset{} and \ssp{} addresses this. Indeed, this
+is not an issue with our implementation, but a more fundamental issue with our
+analysis which was found by it. In that sense, we can see that the benefits of
+computationally encoding the analysis are already emerging.
 
-An additional point to note is that, when evaluating candidates,
-instead of considering equally-ranked constraints individually, we
-treat violations between them cumulatively. As a result, since
-\ssp{} generates multiple violations at a time (for instance [i]
-will generate 3 violations, per Figure
-\pref{table:vowel-sonority-v2}), in order for \maxc{} to
-adequately regulate against deletion, it must generate multiple
-(specifically, 3) violations for each deletion. In addition,
-\onset{}, being equally ranked to \ssp{} and \maxc{}, must also
-do this. This quirk of our implementation is certainly
-unconventional among the OT literature, and further study might
-better support some of these choices (or provide more elegant
-alternative solutions).
+An additional point to note is that, when evaluating candidates, instead of
+considering equally-ranked constraints individually, we treat violations
+between them cumulatively. As a result, since \ssp{} generates multiple
+violations at a time (for instance [i] will generate 3 violations, per Figure
+\pref{table:vowel-sonority-v2}), in order for \maxc{} to adequately regulate
+against deletion, it must generate multiple (specifically, 3) violations for
+each deletion. In addition, \onset{}, being equally ranked to \ssp{} and
+\maxc{}, must also do this. This quirk of our implementation is certainly
+unconventional among the OT literature, and further study might better support
+some of these choices (or provide more elegant alternative solutions).
 
-Beyond these issue, our analysis is sound. Given an underlying
-Yoruba form, the program implementing it will produce the
-appropriately vowel-elided surface form. In other words, given
-the information in the `Underlying' columns of tables
-\pref{fig:deletion-simple}, \pref{fig:deletion-except-1},
-\pref{fig:deletion-except-1}, our program outputs the
-form expected in the 'Surface' columns (excepting
-\textipa{/n\'i-oko/} \rightarrow{} \textipa{[l\'oko]},
-since the n \rightarrow{} l transition occurs for reasons
-unrelated to our analysis; we also ignore tone markers
-here, because it is a related but orthogonal phenomenon
-unhandled by our original analysis).
+Beyond these issue, our analysis is sound. Given an underlying Yoruba form, the
+program implementing it will produce the appropriately vowel-elided surface
+form. In other words, given the information in the `Underlying' columns of
+tables \pref{fig:deletion-simple}, \pref{fig:deletion-except-1},
+\pref{fig:deletion-except-1}, our program outputs the form expected in the
+'Surface' columns (excepting \textipa{/n\'i-oko/} \rightarrow{}
+\textipa{[l\'oko]}, since the n \rightarrow{} l transition occurs for reasons
+unrelated to our analysis; we also ignore tone markers here, because it is a
+related but orthogonal phenomenon unhandled by our original analysis).
 
-In fact, it will even produce the intuitively expected surface
-form when given underlying forms which do not exist. For
-instance, given \textipa{[so-el\'u]} as an input, it will produce
-\textipa{/sel\'u/}, due to \maxplusv{} regulating more
-strictly against initial vowel deletion than deletion of
-final vowels.
+In fact, it will even produce the intuitively expected surface form when given
+underlying forms which do not exist. For instance, given \textipa{[so-el\'u]}
+as an input, it will produce \textipa{/sel\'u/}, due to \maxplusv{} regulating
+more strictly against initial vowel deletion than deletion of final vowels.
 
@@
 
-In this paper, we presented a classic Optimality-Theoretic
-analysis of conjoined vowel elision in Yoruba. We then
-implemented it as a computer program, finding a subtle
-issue in the analysis. With these issues addressed in the
-implementation, our program outputs the correctly
-vowel-elided surface form, given an underlying
-representation. Future work could go in several interesting
-directions. Within the domain of Yoruba, the original
-analysis and its implementation could be extended to handle
-the assimilatory strategy for dealing with sequences of
-vowels at morpheme boundaries (as discussed in
-\cite{ola2002yoruba}), as well as tone. This would likely
-involve extending surface candidate generation to the full
-idealized algorithm described in Section
-\ref{subsec:approach}, contending with epenthesis,
-metathesis, and segment changes in addition to deletion.
-The implementation, as it exists, faithfully models the
-parts of the algorithm in \ref{subsec:approach} that are
-necessary in the context of Yoruba, so this would largely
-be a case of extending the existing machinery.
+In this paper, we presented a classic Optimality-Theoretic analysis of
+conjoined vowel elision in Yoruba. We then implemented it as a computer
+program, finding a subtle issue in the analysis. With these issues addressed in
+the implementation, our program outputs the correctly vowel-elided surface
+form, given an underlying representation. Future work could go in several
+interesting directions. Within the domain of Yoruba, the original analysis and
+its implementation could be extended to handle the assimilatory strategy for
+dealing with sequences of vowels at morpheme boundaries (as discussed in
+\cite{ola2002yoruba}), as well as tone. This would likely involve extending
+surface candidate generation to the full idealized algorithm described in
+Section \ref{subsec:approach}, contending with epenthesis, metathesis, and
+segment changes in addition to deletion. The implementation, as it exists,
+faithfully models the parts of the algorithm in \ref{subsec:approach} that are
+necessary in the context of Yoruba, so this would largely be a case of
+extending the existing machinery.
 

### 2023-05-08 paper.tex
Spruce up conclusion a bit

@@
 syllable structure constraints). This text will specifically explore the
-second strategy of deletion, and in doing so, will specially attempt to provide
+second strategy of deletion (in particular, ignoring tone markers as a related
+but orthogonal case), and in doing so, will specially attempt to provide
 an analysis that is simple enough to be feasible to verify computationally. That
@@
 
-\subsection{Analysis on corpus data}
+\subsection{Performance Analysis on Corpus Data}
 
@@
 conjoined vowel elision in Yoruba. We then implemented it as a computer
-program, finding a subtle issue in the analysis. With these issues addressed in
+program, finding a subtle issue in the analysis in doing so. With these issues
+addressed in
 the implementation, our program outputs the correctly vowel-elided surface
@@
 necessary in the context of Yoruba, so this would largely be a case of
-extending the existing machinery.
+extending the existing machinery. To conclude, we have found that encoding our
+conventional Optimality-Theoretic analysis computationally was highly effective
+at identifying weaknesses and oversights, and at affirming its soundness after
+addressing them. Optimality Theory lends itself quite naturally to computational
+interpretations, and given our encouraging results, we look forward to the rich
+body of work that lies ahead.
 

### 2023-05-08 paper.tex
Address line breaking

@@
 program, finding a subtle issue in the analysis in doing so. With these issues
-addressed in
-the implementation, our program outputs the correctly vowel-elided surface
-form, given an underlying representation. Future work could go in several
-interesting directions. Within the domain of Yoruba, the original analysis and
-its implementation could be extended to handle the assimilatory strategy for
-dealing with sequences of vowels at morpheme boundaries (as discussed in
-\cite{ola2002yoruba}), as well as tone. This would likely involve extending
-surface candidate generation to the full idealized algorithm described in
-Section \ref{subsec:approach}, contending with epenthesis, metathesis, and
-segment changes in addition to deletion. The implementation, as it exists,
-faithfully models the parts of the algorithm in \ref{subsec:approach} that are
-necessary in the context of Yoruba, so this would largely be a case of
-extending the existing machinery. To conclude, we have found that encoding our
-conventional Optimality-Theoretic analysis computationally was highly effective
-at identifying weaknesses and oversights, and at affirming its soundness after
-addressing them. Optimality Theory lends itself quite naturally to computational
-interpretations, and given our encouraging results, we look forward to the rich
-body of work that lies ahead.
+addressed in the implementation, our program outputs the correctly vowel-elided
+surface form, given an underlying representation. Future work could go in
+several interesting directions. Within the domain of Yoruba, the original
+analysis and its implementation could be extended to handle the assimilatory
+strategy for dealing with sequences of vowels at morpheme boundaries (as
+discussed in \cite{ola2002yoruba}), as well as tone. This would likely involve
+extending surface candidate generation to the full idealized algorithm
+described in Section \ref{subsec:approach}, contending with epenthesis,
+metathesis, and segment changes in addition to deletion. The implementation, as
+it exists, faithfully models the parts of the algorithm in
+\ref{subsec:approach} that are necessary in the context of Yoruba, so this
+would largely be a case of extending the existing machinery. To conclude, we
+have found that encoding our conventional Optimality-Theoretic analysis
+computationally was highly effective at identifying weaknesses and oversights,
+and at affirming its soundness after addressing them. Optimality Theory lends
+itself quite naturally to computational interpretations, and given our
+encouraging results, we look forward to the rich body of work that lies ahead.
 
